---
title: "RPS_Exeriment"
format: html
editor: visual
---

## Introduction

ELO is a popular rating system to use an IRT model to rate players, but includes a major transitivity assumption. $P(win | \theta_i, \theta_j)$

## Rock-Paper-Scissors IRT Experiment

```{r}
library(tidyverse)
library(future.apply)
library(dplyr)
library(ggplot2)
library(patchwork)
```

## Simple ELO model

```{r}
## Simulating the game data
N_ids <- 100
N_games <- 5000
elo_thetas <- rnorm(N_ids)

simulate_game <- function(thetas){
  t_ids <- sample(c(1:length(thetas)),2,replace=FALSE)
  t <- thetas[t_ids]
  dt <- t[1]-t[2]
  p <- 1/(1+exp(-dt))
  w <- rbinom(n=1,size=1,prob=p)
  res <- c(t_ids[1],t_ids[2],w)
  return(res)
}


elo_matches <- as.data.frame(do.call(rbind, replicate(N_games,simulate_game(elo_thetas),simplify = FALSE)))

elo_matches <- elo_matches |> rename(player1="V1",
                             player2="V2",
                             result="V3")
```

### Fitting ELO Model

```{r}
LogLikMatch <- function(data, theta) {
  p <- plogis(theta[data$player1] - theta[data$player2])
  ll <- sum(data$result * log(p) + (1 - data$result) * log(1 - p))
  return(ll)
}

starting_theta <- rnorm(N_ids)

sol <- optim(
  starting_theta,
  LogLikMatch,
  data = elo_matches,
  method = 'L-BFGS-B',
  lower = -3,
  upper = 3,
  control = list(fnscale = -1, maxit = 1e6)
)

cor(sol$par, elo_thetas)
plot(sol$par, elo_thetas)

```

## Simulating RPS

```{r}
## Simulating the game data
set.seed(1729)
N_ids <- 100
N_games <- 1000
N_opts <- 3 ## For rock-paper-scissors
rps_thetas <- rep(0,N_ids)

rps_winner <- function(x,y,N_opts){
  a <- min(x,y)
  b <- max(x,y)
  if(a==1 & b==N_opts){
    return(a)
  }else{
    return(b)
  }
}

simulate_rps <- function(thetas,N_opts){
  t_ids <- sample(c(1:length(thetas)),2,replace=FALSE)
  h <- sample(c(1:N_opts),2,replace=FALSE)
  h_win <- rps_winner(h[1],h[2],N_opts)
  w <- ifelse(h_win==h[1],1,0)
  res <- c(t_ids[1],t_ids[2],h[1],h[2],w)
  return(res)
}


rps_matches <- as.data.frame(do.call(rbind, replicate(N_games,simulate_rps(rps_thetas,N_opts),simplify = FALSE)))

rps_matches <- rps_matches |> rename(player1="V1",
                             player2="V2",
                             hand1 = "V3",
                             hand2 = "V4",
                             result="V5")
```

```{r}
LogLik_RPS <- function(data, params, N_ids) {
  theta <- params[1:N_ids]
  phi <- c(0,params[(N_ids+1):length(params)])
  d_theta <- theta[data$player1] - theta[data$player2]
  # forcing d_theta to be 0
  # d_theta <- 0
  d_phi <- phi[data$hand1] - phi[data$hand2]
  c <- (1+d_theta)*(1+2*atanh(sinpi(d_phi)))-1
  # c <- (1+d_theta)*(1+log((1+sinpi(d_phi))/(1-sinpi(d_phi))))
  p <- plogis(c)
  ll <- sum(data$result * log(p) + (1 - data$result) * log(1 - p))
  return(ll)
}

estimate_nt_elo <- function(N_ids,N_opts,rps_matches,starting_phi){
  starting_theta <- rnorm(N_ids)
  starting_phi <- starting_phi
  
  # starting_phi <- (c(1:N_opts)-1)*2/N_opts
  # starting_phi <- sample(starting_phi,N_opts,FALSE)
  
  starting_params <- c(starting_theta,starting_phi)
  
  sol <- optim(
    starting_params,
    LogLik_RPS,
    data = rps_matches,
    N_ids = N_ids,
    method = 'CG',
    # lower = -3,
    # upper = 3,
    control = list(fnscale = -1,
                   maxit = 1e6
  ))
  
  return(list(sol=sol,starting_phi=starting_phi))
}

starting_phi <- runif(N_opts-1,min=0,max=2)
res <- estimate_nt_elo(N_ids, N_opts, rps_matches,starting_phi)
sol <- res$sol
starting_phi <- res$starting_phi


```

### Visualizing the results

```{r}
est_theta <- as.data.frame(list(est=sol$par[1:N_ids],
                  true=rps_thetas))
est_phi <- as.data.frame(list(est=c(0,sol$par[(N_ids+1):length(sol$par)]),
                r=rep(1,N_opts),
                type=rep("est",N_opts),
                id=c(1:N_opts)))
start_phi <- as.data.frame(list(est=c(0,starting_phi),
                r=rep(1,N_opts),
                type=rep("init",N_opts),
                id=c(1:N_opts)))

est_phi <- rbind(est_phi,start_phi)

est_phi <- est_phi |> mutate(phi=pi*(est%%2))

est_theta_mu <- round(mean(est_theta$est),2)
est_theta_var <- round(sd(est_theta$est)^2,3)

p_phi <- ggplot(est_phi) +
  geom_point(aes(r, phi,color=factor(id),shape=factor(type))) + 
  coord_radial(theta="y",
               start=pi/2,
               direction=-1,
               rlim=c(0,1),
               thetalim = c(0,2*pi)) +
  ggtitle("Est. Phi") +
  labs(color = "Hand",shape="") +
  scale_color_discrete(labels = c("Rock", "Paper", "Scissors"))

p_theta <- ggplot(est_theta,aes(x=true,y=est)) + 
  geom_point() + 
  theme_bw() +
  ggtitle(paste0("Est. Theta ~ N(",est_theta_mu,",",est_theta_var,")"))

p_theta | p_phi
```

#### Checking model fit stability

Simulating 1000 times to check how often the model fit fails or returns degenerate results. Takes some time to run.

```{r}

plan(multisession, workers = parallel::detectCores() - 1)

set.seed(1729)
N_sim <- 1e2
N_ids <- 100
N_games <- 1000
N_opts <- 3 

run_single_sim <- function(i, N_ids, N_games, N_opts) {

  rps_thetas <- rep(0, N_ids)
  
  rps_matches <- as.data.frame(do.call(rbind, 
                                       replicate(N_games,
                                                 simulate_rps(rps_thetas, N_opts),
                                                 simplify = FALSE)))
  
  colnames(rps_matches) <- c("player1", "player2", "hand1", "hand2", "result")
  
  starting_phi <- runif(N_opts-1,min=0,max=2)
  sol <- try(estimate_nt_elo(N_ids, N_opts, rps_matches, starting_phi)$sol, silent = TRUE)
  
  if (inherits(sol, "try-error")) {
    return(list(phi = NULL, theta = NULL, starting_phi=starting_phi))
  } else {
    # Process phi
    est_phi <- data.frame(
      est = c(0, sol$par[(N_ids + 1):length(sol$par)]),
      r = 1,
      id = 1:N_opts
    ) %>% mutate(phi = pi * (est %% 2))
    
    # Process theta
    est_theta_pars <- data.frame(
      mean = mean(sol$par[1:N_ids]),
      var = var(sol$par[1:N_ids])
    )
    
    return(list(phi = est_phi, theta = est_theta_pars, starting_phi=starting_phi))
  }
}


results <- future_lapply(1:N_sim, run_single_sim, 
                         N_ids = N_ids, N_games = N_games, N_opts = N_opts, 
                         future.seed = TRUE)

check_res <- function(res){
  return(is.null(res$phi))
}

successful_results <- results[!sapply(results, check_res)]
failures <- N_sim - length(successful_results)

est_phi_sim <- bind_rows(lapply(successful_results, `[[`, "phi"))
est_theta_sim <- bind_rows(lapply(successful_results, `[[`, "theta"))
starting_phi_sim <- c(lapply(results, `[[`, "starting_phi"))

print(paste0("% optimization fails: ", round(100 * failures / N_sim, 2)))
```

```{r}
p_phi <- ggplot(est_phi_sim) +
  geom_point(aes(r, phi,color=factor(id)),alpha=0.5) + 
  coord_radial(theta="y",
               start=pi/2,
               direction=-1,
               rlim=c(0,1),
               thetalim = c(0,2*pi)) +
  ggtitle("Est. Phis") +
  labs(color = "Hand") +
  scale_color_discrete(labels = c("Rock", "Paper", "Scissors"))

p_theta <- ggplot(est_theta_sim,aes(x=mean,y=var)) + 
  geom_point() + 
  theme_bw() +
  ggtitle(paste0("Est. Thetas ~ N(Mean,Var)"))

p_theta | p_phi
```

Investigating cases with failures

-   Bad starting conditions: Seems to work for a different simulation of matches with these starting params

-   Cases we get poor Phi estimates

```{r}
failed_fit_ids <- sapply(results, check_res)
failed_starting_phis <- starting_phi_sim[failed_fit_ids]

for(i in c(1:length(failed_starting_phis))){
  res <- try(estimate_nt_elo(N_ids, N_opts, rps_matches,failed_starting_phis[[i]]),
             silent = TRUE)  
  print(res)
}



```

Check optim output for weird cases for convergence warnings

Start at the "right answer" and see if things still diverge

## Simulating RPS with individual "luck" (ability)

```{r}
## Simulating the game data
set.seed(1729)
N_ids <- 100
N_games <- 5000
N_opts <- 3 ## For rock-paper-scissors
rps_luck_thetas <- rnorm(N_ids)

rps_winner <- function(x,y,N_opts){
  a <- min(x,y)
  b <- max(x,y)
  if(a==1 & b==N_opts){
    return(a)
  }else{
    return(b)
  }
}

simulate_rps_luck <- function(thetas,N_opts){
  t_ids <- sample(c(1:length(thetas)),2,replace=FALSE)
  rps_wins <- list(c(2,1),c(3,2),c(1,3))
  h <- sample(rps_wins,1)[[1]]
  ## Luckier player to win
  dt <- thetas[t_ids[1]] - thetas[t_ids[2]]
  w <- rbinom(1,1,plogis(dt))
  if(w==0){
    h <- c(h[2],h[1])
  }
  res <- c(t_ids[1],t_ids[2],h[1],h[2],w)
  return(res)
}


rps_matches <- as.data.frame(do.call(rbind, replicate(N_games,simulate_rps_luck(rps_luck_thetas,N_opts),simplify = FALSE)))

rps_matches <- rps_matches |> rename(player1="V1",
                             player2="V2",
                             hand1 = "V3",
                             hand2 = "V4",
                             result="V5")
```

```{r}
starting_theta <- rnorm(N_ids)
# starting_phi <- runif(N_opts-1,min=0,max=2)
starting_phi <- rep(0.25,N_opts-1)



starting_params <- c(starting_theta,starting_phi)

sol <- optim(
  starting_params,
  LogLik_RPS,
  data = rps_matches,
  N_ids = N_ids,
  method = 'CG',
  # lower = -3,
  # upper = 3,
  control = list(fnscale = -1,
                 maxit = 1e6
))

est_theta <- as.data.frame(list(est=sol$par[1:N_ids],
                  true=rps_luck_thetas))
est_phi <- as.data.frame(list(est=c(0,sol$par[(N_ids+1):length(sol$par)]),
                r=rep(1,N_opts),
                type=rep("est",N_opts),
                id=c(1:N_opts)))
start_phi <- as.data.frame(list(est=c(0,starting_phi),
                r=rep(1,N_opts),
                type=rep("init",N_opts),
                id=c(1:N_opts)))

est_phi <- rbind(est_phi,start_phi)

est_phi <- est_phi |> mutate(phi=pi*(est%%2))

est_theta_mu <- round(mean(est_theta$est),2)
est_theta_var <- round(var(est_theta$est),3)

p_phi <- ggplot(est_phi) +
  geom_point(aes(r, phi,color=factor(id),shape=factor(type))) + 
  coord_radial(theta="y",
               start=pi/2,
               direction=-1,
               rlim=c(0,1),
               thetalim = c(0,2*pi)) +
  ggtitle("Est. Phi") +
  labs(color = "Hand",shape="") +
  scale_color_discrete(labels = c("Rock", "Paper", "Scissors"))

r_value <- round(cor(est_theta$est,est_theta$true),2)
p_theta <- ggplot(est_theta,aes(x=true,y=est)) + 
  geom_point() + 
  annotate("text", x = 2, y = -0.2, label = paste("r =", r_value)) +
  theme_bw() +
  ggtitle(paste0("Est. Theta ~ N(",est_theta_mu,",",est_theta_var,")"))

p_theta | p_phi
```

Simulating 1000 times to check model fit stability. Takes a long time to run.

```{r}

plan(multisession, workers = parallel::detectCores() - 1)

set.seed(1729)
N_sim <- 1e4
N_ids <- 100
N_games <- 1000
N_opts <- 3 


run_single_sim <- function(i, N_ids, N_games, N_opts) {
  
  rps_luck_thetas <- rnorm(N_ids)
  
  rps_matches <- as.data.frame(do.call(rbind, 
                                       replicate(N_games,
                                                 simulate_rps_luck(rps_luck_thetas, N_opts),
                                                 simplify = FALSE)))
  
  colnames(rps_matches) <- c("player1", "player2", "hand1", "hand2", "result")
  
  sol <- try(estimate_nt_elo(N_ids, N_opts, rps_matches)$sol
             , silent = TRUE)
  
  if (inherits(sol, "try-error")) {
    return(NULL) # Indicate a failure
  } else {
    # Process phi
    est_phi <- data.frame(
      est = c(0, sol$par[(N_ids + 1):length(sol$par)]),
      r = 1,
      id = 1:N_opts
    ) %>% mutate(phi = pi * (est %% 2))
    
    # Process theta
    est_theta_pars <- data.frame(
      mean = mean(sol$par[1:N_ids]),
      var = var(sol$par[1:N_ids]),
      cor = cor(sol$par[1:N_ids],rps_luck_thetas)
    )
    
    return(list(phi = est_phi, theta = est_theta_pars))
  }
}

results <- future_lapply(1:N_sim, run_single_sim, 
                         N_ids = N_ids, N_games = N_games, N_opts = N_opts, 
                         future.seed = TRUE)

successful_results_luck <- results[!sapply(results, is.null)]
failures_luck <- N_sim - length(successful_results)

est_phi_sim_luck <- bind_rows(lapply(successful_results_luck, `[[`, "phi"))
est_theta_sim_luck <- bind_rows(lapply(successful_results_luck, `[[`, "theta"))

print(paste0("% optimization fails: ", round(100 * failures_luck / N_sim, 2)))
```

```{r}
p_phi <- ggplot(est_phi_sim_luck) +
  geom_point(aes(r, phi,color=factor(id)),alpha=0.5) + 
  coord_radial(theta="y",
               start=pi/2,
               direction=-1,
               rlim=c(0,1),
               thetalim = c(0,2*pi)) +
  ggtitle("Est. Phis") +
  labs(color = "Hand") +
  scale_color_discrete(labels = c("Rock", "Paper", "Scissors"))

p_theta <- ggplot(est_theta_sim_luck,aes(x=mean,y=var)) + 
  geom_point() + 
  theme_bw() +
  ggtitle(paste0("Est. Thetas ~ N(Mean,Var)"))

p_theta | p_phi
```

## Generalized case

Each player employs a different strategy (pick RPS with varying probabilities) facing each other in a game

```{r}
## Simulating the game data
set.seed(1729)
N_ids <- 5e2
N_games <- 1e4
N_opts <- 3 ## For rock-paper-scissors
rps_thetas <- list()

for(i in c(1:N_ids)){
  theta <- runif(3)
  theta <- theta/sum(theta)
  rps_thetas[[i]] <- theta
}

rps_winner <- function(x,y,N_opts){
  a <- min(x,y)
  b <- max(x,y)
  if(a==1 & b==N_opts){
    return(a)
  }else{
    return(b)
  }
}

simulate_rps <- function(thetas,N_opts){
  t_ids <- sample(c(1:length(thetas)),2,replace=FALSE)
  h1 <- 0
  h2 <- 0
  while(h1 == h2){
    h1 <- sample(x=c(1:N_opts),size=1,prob=thetas[[t_ids[1]]]) #player 1 hand 
    h2 <- sample(x=c(1:N_opts),size=1,prob=thetas[[t_ids[2]]]) #player 2 hand   
  }
  
  h_win <- rps_winner(h1,h2,N_opts)
  w <- ifelse(h_win==h1,1,0)
  res <- c(t_ids[1],t_ids[2],h1,h2,w)
  return(res)
}


rps_matches <- as.data.frame(do.call(rbind, replicate(N_games,simulate_rps(rps_thetas,N_opts),simplify = FALSE)))

rps_matches <- rps_matches |> rename(player1="V1",
                             player2="V2",
                             hand1 = "V3",
                             hand2 = "V4",
                             result="V5")
```

```{r}
starting_phi <- rep(0.25,N_opts-1)
res <- estimate_nt_elo(N_ids, N_opts, rps_matches,starting_phi)
sol <- res$sol
starting_phi <- res$starting_phi


```

```{r}

theta_rad <- sapply(rps_thetas, max)
theta_rad_id <- sapply(rps_thetas, which.max)
est_theta <- as.data.frame(list(est=sol$par[1:N_ids],
                  true=theta_rad,
                  true_hand_favored=theta_rad_id,
                  id=c(1:N_ids)))
est_phi <- as.data.frame(list(est=c(0,sol$par[(N_ids+1):length(sol$par)]),
                r=rep(1,N_opts),
                type=rep("est",N_opts),
                id=c(1:N_opts))) |> 
  mutate(phi = pi*(est%%2),
         x = cos(phi),
         y = sin(phi))

```

```{r}

player_stats <- rbind(rps_matches |> select(player1,hand1) |> 
                        rename_with(~ str_remove(., "\\d+$")),
                      rps_matches |> select(player2,hand2) |> 
                        rename_with(~ str_remove(., "\\d+$")))

player_stats <- player_stats |> 
  left_join(est_theta,
            by=join_by(player==id)) |>
  left_join(est_phi |> filter(type=="est") |> select(id,phi,x,y),
            by=join_by(hand==id)) |>
  rename(theta = est)


player_pos <- player_stats |> 
  group_by(player) |> 
  summarise(across(c(x,y,true_hand_favored,true),mean)) |>
  mutate(r=sqrt(x^2+y^2),
         phi=atan2(y,x)) |>
  rename(favored_hand=true_hand_favored)
  


p_phi_radial <- ggplot(player_pos) +
  geom_point(aes(r, phi,color=factor(favored_hand))) +
  coord_radial(theta="y",
               start=pi/2,
               direction=-1,
               rlim=c(0,1),
               thetalim = c(0,2*pi)) +
  ggtitle("Player estimates") +
  labs(color = "Favored Hand") +
  scale_color_discrete(labels = c("Rock", "Paper", "Scissors"))

r_value <- cor(player_pos$r,player_pos$true)
p_phi_cart <- ggplot(player_pos) +
  geom_point(aes(true, r)) + 
  annotate("text", x = 0.9, y = 0, label = paste("r =", round(r_value,3))) +
  ggtitle("Player ability/luck estimates")

p_phi_cart | p_phi_radial
```