---
title: "NBA Non Transitive Model"
author: "Kaushik Mohan"
date: "2026-09-10"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(Matrix)
library(see)
library(ggrepel)
```

NBA 2025-26 game outcome data from https://www.basketball-reference.com/leagues/NBA_2026_games.html


```{r}
games <- read.csv("~/Documents/PhD/Spring 26/Measurement/NonTransitiveELO/NBA_25_26.csv")
teams <- sort(unique(c(games$Home.Neutral, games$Visitor.Neutral)))

games_clean <- games |> filter(Notes != "NBA Cup") |>
  mutate(Date = mdy(Date)) |>
  filter(Date<="2026-04-12") |>
  select(Home.Neutral, Visitor.Neutral, PTS.1, PTS) |> 
  mutate(result = ifelse(PTS.1 > PTS,1,0)) |>
  select(Home.Neutral, Visitor.Neutral, result) |>
  rename(player1=Home.Neutral,
         player2=Visitor.Neutral) |>
  mutate(player1 = match(player1,teams),
         player2 = match(player2,teams))

N_ids <- N_opts <- length(teams)

```


```{r}
season_matrix <- read.csv("~/Documents/PhD/Spring 26/Measurement/NonTransitiveELO/nba_25_26_matrix.csv") 
season_matrix <- season_matrix |> 
  select(-Rk) |>
  mutate(Team = colnames(season_matrix)[2:31]) |>
  pivot_longer(cols=2:31,values_to = "record",names_to="team2") |>
  mutate(record = case_when(
    record=="" ~ "0-1",
    TRUE ~ record
  )) |>
  rowwise() |>
  mutate(p = as.numeric(str_split(record,"-")[[1]])[1]/sum(as.numeric(str_split(record,"-")[[1]]))) |>
  select(-record) |>
  pivot_wider(names_from = team2,values_from = p)

rownames(season_matrix) <- season_matrix$Team

season_matrix <- season_matrix |> select(-Team) |> as.matrix()

```



```{r}
games_mat_long <- games_clean |> group_by(player1, player2) |> summarize(p=mean(result),.groups="drop")
payoff_mat <- games_mat_long |>
  pivot_wider(names_from = player2, values_from = p, values_fill = 0) |>
  select(-player1) |>
  as.matrix() 

payoff_mat <- payoff_mat[c(1:30),as.character(sort(as.numeric(colnames(payoff_mat))))]
K_skew <- (payoff_mat - t(payoff_mat)) / 2
```


```{r}
payoff_matrix <- function(N,eps=0) {
  eps <- abs(eps)
  # generate square matrix with 0s
  
  A <- matrix(0, nrow = N, ncol = N)
  
  # Fill upper triangle with payoff
  for(i in c(1:(N-1))){
    A[i,(i+1):N] <- rep(c(-1+eps, 1-eps), length.out = (N-i))
  }
  
  # Enforce skew-symmetry for cyclic game
  A <- (A - t(A))/2
  
  # Convert to Probabilities
  pA <- (A + 0.5)
  return(pA)
}

```



```{r}
P <- payoff_matrix(3,0.5)
P
round(Schur(qlogis(P))$Q,2)
round(Schur(qlogis(P))$T,2)
a <- round(Schur(qlogis(P))$Q,2)[1,1:2]
b <- round(Schur(qlogis(P))$Q,2)[2,1:2]
dp <- sum(a*b)
norm_a <- sqrt(sum(a^2))
norm_b <- sqrt(sum(b^2))
acos(dp/(norm_a*norm_b))*180/pi

```
### Models

```{r}
mat_from_lt <- function(vals,n){
  m <- matrix(0, n,n)
  m[lower.tri(m)] <- vals
  m <- t(m)
  m[lower.tri(m)] <- -vals
  m <- t(m)
  return(m)
}

LogLik_NT_NBA <- function(params, data, N_ids, N_opts,
                      lambda_t=0.1, lambda_k=0,
                      theta_centering = FALSE,
                      ridge_penalty = FALSE,
                      home_advantage = TRUE) {

  if(home_advantage){
    offset <- 1
  }else{
    offset <- 0
  }
  if (theta_centering) {
    theta <- params[1:N_ids]
    theta <- theta - mean(theta)
    phis <- params[(N_ids + 1):(length(params)-offset)]
  } else {
    theta <- c(0, params[1:(N_ids - 1)])
    phis <- params[N_ids:(length(params)-offset)]
  }
  
  
  # Player skill difference
  d_theta <- theta[data$player1] - theta[data$player2]

  # Setting up the log-odds payoff matrix
  K_mat <- mat_from_lt(phis, N_opts)
  k_ab <- K_mat[cbind(data$player1, data$player2)]

  if(home_advantage){
    # Home-field advantage
    d_h <- tail(params,1)  
  }else{
    d_h <- 0
  }
  
  # Additive logit predictor
  logit <- d_h + d_theta + k_ab 
  p <- plogis(logit)
  # p <- pmin(1 - 1e-6, pmax(1e-6, p))

  # Negative log-likelihood with Ridge penalty on theta
  ll <- sum(data$result * log(p) + (1 - data$result) * log(1 - p))
  ## Lasso penalty
  penalty <- ifelse(ridge_penalty, lambda_t * sum(theta^2) + lambda_k*sum(abs(phis)), 0)
  neg_ll <- -ll + penalty

  if (is.finite(neg_ll)) neg_ll else 1e10
}

# Estimate ELO + Non-Transitive Payoffs
estimate_nt_elo_nba <- function(N_ids, N_opts, matches,
                            lambda_t=0.1,
                            lambda_k=0.1,
                            theta_centering = TRUE,
                            ridge_penalty = TRUE,
                            home_advantage=TRUE) {

  n_theta <- ifelse(theta_centering, N_ids, N_ids - 1)
  n_payoff <- N_opts * (N_opts - 1) / 2

  # Initialize near zero in log-odds space
  starting_theta <- rnorm(n_theta, mean = 0, sd = 0.1)
  starting_po <- rnorm(n_payoff, mean = 0, sd = 0.1)
  if(home_advantage){
    starting_home_adv <- rnorm(1, mean = 0, sd = 0.1)
    starting_params <- c(starting_theta, starting_po, starting_home_adv)  
  }else{
    starting_params <- c(starting_theta, starting_po)  
  }
  

  # Unconstrained bounds for log-odds payoffs
  lower_bounds <- rep(-Inf, n_theta + n_payoff)
  upper_bounds <- rep(Inf, n_theta + n_payoff)

  sol <- optim(
    par = starting_params,
    fn = LogLik_NT_NBA,
    data = matches,
    N_ids = N_ids,
    N_opts = N_opts,
    lambda_t=lambda_t,
    lambda_k=lambda_k,
    theta_centering = theta_centering,
    ridge_penalty = ridge_penalty,
    home_advantage = home_advantage,
    method = 'L-BFGS-B',
    hessian = FALSE,
    lower = lower_bounds,
    upper = upper_bounds,
    control = list(fnscale = 1, maxit = 1e6)
  )

  return(sol)
}

LogLikELO <- function(data, theta) {
  d_h <- tail(theta,1)
  p <- plogis(d_h + theta[data$player1] - theta[data$player2])
  ll <- sum(data$result * log(p) + (1 - data$result) * log(1 - p))
  return(ll)
}

fit_elo_model <- function(N_players,matches){
  
  starting_theta <- rnorm(N_players+1)

  sol <- optim(
    starting_theta,
    LogLikELO,
    data = matches,
    method = 'L-BFGS-B',
    lower = -3,
    upper = 3,
    control = list(fnscale = -1, maxit = 1e6)
  )

  return(sol)
}
```


## Viz

```{r}
plot_heatmap <- function(mat, labels = NULL) {
  labels <- str_split_i(labels," ",-1)
  # Apply row labels if provided
  if (!is.null(labels)) {
    rownames(mat) <- make.unique(as.character(labels))
    colnames(mat) <- make.unique(as.character(labels))
  } else if (is.null(rownames(mat))) {
    rownames(mat) <- paste0("Row_", seq_len(nrow(mat)))
    colnames(mat) <- paste0("Col_", seq_len(ncol(mat)))
  }
  
  # Reshape matrix into long format for ggplot
  df_long <- as.data.frame(mat) %>%
    mutate(Row = factor(rownames(.), levels = rev(rownames(.)))) %>%
    pivot_longer(cols = -Row, names_to = "Column", values_to = "Value") %>%
    mutate(Column = factor(Column, levels = colnames(mat)))
  
  # Build ggplot heatmap
  ggplot(df_long, aes(x = Column, y = Row, fill = Value)) +
    geom_tile(color = "white", linewidth = 0) +
    # scale_fill_material(discrete = FALSE) + 
    scale_fill_gradient2(low="#D55E00",
                         # mid="#999999",
                         high="#009E73",
                         midpoint=mean(df_long$Value)
                         ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      panel.grid = element_blank(),
      axis.title = element_blank()
    ) +
    coord_fixed()
}
```

## Season match-up win probabilities

```{r}
plot_heatmap(season_matrix, teams)
```

## Simple ELO model (with home advantage) 

```{r}
nba_elo_model <- fit_elo_model(N_ids,games_clean)
team_ratings <- data.frame(team=teams,elo_theta=nba_elo_model$par[1:N_ids])
d_hf_elo <- tail(nba_elo_model$par,1)
```


## Non-transitive ELO model (with home advantage)

### Tuning penalty param

```{r,eval=FALSE}
lambda_t <- c(0,0.1,seq(0.2,0.3,0.01),0.4,0.5,0.75,1,2,5)
lambda_k <- c(0,0.5,seq(0.75,1.25,0.05),1.5,1.75,2,3,4,5)
lambdas <- expand.grid(lambda_t, lambda_k)

nlls <- rep(NA,nrow(lambdas))
ks <- rep(NA,nrow(lambdas))

for(i in c(1:nrow(lambdas))){
  m <- estimate_nt_elo_nba(N_ids,N_opts,games_clean,lambda_t = lambdas$Var1[i],lambda_k = lambdas$Var2[i])
  ## Slightly different definition of "free" parameters for BIC where I'm using just the estimated parameters after lasso and ignoring the ones that shrink to 0
  ks[i] <- sum(round(m$par,2) != 0)
  nlls[i] <- m$value
}

reg_stats <- data.frame(lambda_theta=lambdas$Var1,
                        lambda_k=lambdas$Var2) |>
  mutate(nll = nlls,
         k = ks,
         aic= 2*nll+2*k,
         bic = 2*nll +k*log(nrow(games_clean)))

```

## AIC and BIC for different combinations of lambda

```{r}
reg_stats <- read.csv("~/Documents/PhD/Spring 26/Measurement/NonTransitiveELO/reg_stats.csv")
ggplot(reg_stats |> filter(lambda_theta >=0.23,lambda_theta<=0.27),aes(x=lambda_k,y=aic,color=factor(lambda_theta))) + 
  geom_line() + 
  labs(title = "AIC") +
  theme_bw() +
  theme_minimal()

ggplot(reg_stats |> filter(lambda_theta >=0.23,lambda_theta<=0.27),aes(x=lambda_k,y=bic,color=factor(lambda_theta))) + 
  geom_line() + 
  labs(title = "BIC") +
  theme_bw() +
  theme_minimal()

```

## Model fit at optimal lambdas

```{r}
nba_NT_elo_model <- estimate_nt_elo_nba(N_ids,N_opts,games_clean,lambda_t = 0.24, lambda_k = 1)
```


```{r}
team_ratings <- team_ratings |> mutate(nt_elo_theta=nba_NT_elo_model$par[1:N_ids])
n_payoff <- N_opts * (N_opts - 1) / 2
nt_mat <- mat_from_lt(nba_NT_elo_model$par[(N_ids+1):(N_ids+n_payoff)],N_ids)
d_hf_ntelo <- tail(nba_NT_elo_model$par,1)
```

### Non-transitive component of fit and decomposition

Heavily regularized model leaves us with just one dominant non-transitive cycle between the Hawks, Bulls, and Pacers. 
Pacers were one one of the worst teams (#29 of 30) in the league last season (injury to their main player), but won all 4 of their games against the Bulls (#22 of 30).
Bulls on the other hand won all 3 of their games against the Hawks (#11 of 30), though they are the worse team on paper (from league standings, estimated ELO).
Given the vast skill differential between the Hawks and Pacers (#11 vs. #29), the Hawks beat the pacers in 2 out of the 3 games played between them. 

It is interesting to see this cycle as it is probably the simplest cycle that "corrects" discrepencies in the observed results not explained by the ELO model. 

```{r}
plot_heatmap(nt_mat, teams)
sch_decomp <- Schur(nt_mat)
T_mat <- round(sch_decomp$T,2)
# T_mat
n_cycles <- floor(nrow(nt_mat) / 2)
cycle_strengths <- numeric(n_cycles)
  for (k in 1:n_cycles) {
    idx1 <- 2 * k - 1
    idx2 <- 2 * k
    cycle_strengths[k] <- abs(T_mat[idx1, idx2])
  }

ggplot() + geom_line(aes(x=c(1:n_cycles),y=cycle_strengths)) + labs(x="cycles",y="eigenvalues") + theme_bw() +  theme_minimal() 
# round(sch_decomp$Q,2)[,1:2]
# round(cycle_strengths^2/sum(cycle_strengths^2),2)[1]
```



```{r}
first_cycle_locs <- data.frame(team=teams,
                              x=round(cycle_strengths[1]*sch_decomp$Q,2)[,1],
                              y=round(cycle_strengths[1]*sch_decomp$Q,2)[,2])

first_cycle_locs$dist <- sqrt(first_cycle_locs$x^2 + first_cycle_locs$y^2)

ggplot(first_cycle_locs,aes(x=x,y=y,label=team)) + 
  geom_point(aes(size=dist),color=okabeito_colors(3),show.legend = FALSE) +
  geom_text_repel(size = 4, max.overlaps = 25) +
  theme_bw() +
  theme_minimal() +
  labs(title="Primary Non-transitive Cycle (Dominance: Clockwise)")
  
```

This is 

### LR-test of model

This might be the wrong thing to do since we are comparing likelihood from the ELO model to penalized likelihood from the NT ELO model. I think the right thing to do here is to fit a model without regularization but parameterized based on what we find to be the dominant cycle from the regularization

```{r}
ELO_LR_test <- function(elo_model, nt_elo_model,nt_mat){
  elo_logLik <- elo_model$value
  NT_logLik <- -nt_elo_model$value
  LR_stat <- -2*(elo_logLik - NT_logLik)
  df <- sum(round(nt_mat[upper.tri(nt_mat)],2)!=0)
  return(list(stat=LR_stat, df=df))
}

test <- ELO_LR_test(nba_elo_model, nba_NT_elo_model, nt_mat)
test
pchisq(test$stat, test$df, lower.tail = TRUE)
```

Alternatively, we could use a different statistical test to see if the predicted probabilites of matchups from the two models are different. Rank sum test?

### Comparision of predicted probabilities from the two models

```{r}
games_clean <- games_clean |> mutate(p_elo = plogis(d_hf_elo + 
                                                      nba_elo_model$par[1:N_ids][player1] - nba_elo_model$par[1:N_ids][player2]),
                                     p_nt = plogis(d_hf_ntelo + 
                                                     nba_NT_elo_model$par[1:N_ids][player1] - nba_NT_elo_model$par[1:N_ids][player2] +
                                                     nt_mat[cbind(player1,player2)]))
```


```{r}
ggplot(team_ratings,aes(x=elo_theta,y=nt_elo_theta)) + geom_point() + labs(title="ELO vs. NT-elo theta")
ggplot(games_clean,aes(x=p_elo,y=p_nt,color=factor(result))) + geom_point(size=2,alpha=0.7) + labs(title="Predicted probabilities from ELO and NT-elo models")
ggplot(games_clean, aes(x=p_elo,y=result)) + geom_smooth() + labs(title="predictd prob from ELO model vs. result")
ggplot(games_clean, aes(x=p_nt,y=result)) + geom_smooth() + labs(title="predictd prob from NT-elo model vs. result")
```

### Comparison of estimated "Skill" rating

```{r}
team_ratings_long <- team_ratings |> pivot_longer(cols=2:3,names_to = "model",values_to = "theta")
ggplot(team_ratings_long,aes(x=team,y=theta,color=model)) + 
  geom_point(size=2) + 
  coord_flip() + 
  theme_bw() + 
  theme_minimal() +
  scale_color_see()
```

### Predictive performance on playoff games


```{r}
playoff_games <-games |> filter(Notes != "NBA Cup") |>
  mutate(Date = mdy(Date)) |>
  filter(Date>"2026-04-12") |>
  select(Home.Neutral, Visitor.Neutral, PTS.1, PTS) |> 
  mutate(result = ifelse(PTS.1 > PTS,1,0)) |>
  select(Home.Neutral, Visitor.Neutral, result) |>
  rename(player1=Home.Neutral,
         player2=Visitor.Neutral) |>
  mutate(player1 = match(player1,teams),
         player2 = match(player2,teams))

playoff_games <- playoff_games |> mutate(p_elo = plogis(d_hf_elo + 
                                                      nba_elo_model$par[1:N_ids][player1] - nba_elo_model$par[1:N_ids][player2]),
                                     p_nt = plogis(d_hf_ntelo + 
                                                     nba_NT_elo_model$par[1:N_ids][player1] - nba_NT_elo_model$par[1:N_ids][player2] +
                                                     nt_mat[cbind(player1,player2)]))

playoff_games <- playoff_games |> mutate(logloss_elo = -(result*log(p_elo)+(1-result)*log(1-p_elo)),
                                         logloss_nt = -(result*log(p_nt)+(1-result)*log(1-p_nt)))

print(paste0("Log Loss (ELO): ",round(sum(playoff_games$logloss_elo),2)))
print(paste0("Log Loss (NT): ",round(sum(playoff_games$logloss_nt),2)))
```


