---
title: "Estimating Non-transitive Payoff"
format: pdf
editor: visual
---

## Non-transitive Games

## Payoff Matrix

Non-transitive component of games can be represented by a skew-symmetric payoff matrix. For games where probabilities of a player winning over another is dependent on their individual abilities and a non-transitive payoffs

```{r}
library(tidyverse)
library(future.apply)
library(dplyr)
library(ggplot2)
library(patchwork)
library(latex2exp)
library(see)
options(warn = -1)
```

## Game Simulation functions

```{r}
simulate_elo_game <- function(thetas,N_opts=3){
  t_ids <- sample(c(1:length(thetas)),2,replace=FALSE)
  t <- thetas[t_ids]
  dt <- t[1]-t[2]
  p <- 1/(1+exp(-dt))
  w <- rbinom(n=1,size=1,prob=p)
  ## Hand played is randomly chosen
  h <- sample(c(1:N_opts),2,replace=TRUE)
  res <- c(t_ids[1],t_ids[2],h[1],h[2],w,p)
  return(res)
}

## Non-transitive payoff probability matrix (eps -> [0,2])
payoff_matrix <- function(N,eps=0) {
  eps <- abs(eps)
  # generate square matrix with 0s
  
  A <- matrix(0, nrow = N, ncol = N)
  
  # Fill upper triangle with payoff
  for(i in c(1:(N-1))){
    A[i,(i+1):N] <- rep(c(-1+eps, 1-eps), length.out = (N-i))
  }
  
  # Enforce skew-symmetry for cyclic game
  A <- A - t(A)
  
  # Convert to Probabilities
  pA <- 0.5*(A + 1)
  return(pA)
}


simulate_rps <- function(thetas,N_opts,payoff=NULL,replace=FALSE){
  t_ids <- sample(c(1:length(thetas)),2,replace=FALSE)
  if(is.null(payoff)){
    p <- payoff_matrix(N_opts,eps=0)
  }else{
    p <- payoff
  }
  
  h <- sample(c(1:N_opts),2,replace=replace)
  w <- rbinom(1,1,p[h[1],h[2]])
  # w <- ifelse(prob>0,rbinom(1,1,prob),rbinom(1,1,1+prob))
  res <- c(t_ids[1],t_ids[2],h[1],h[2],w,p[h[1],h[2]])
  return(res)
}

simulate_rps_luck <- function(thetas,N_opts){
  t_ids <- sample(c(1:length(thetas)),2,replace=FALSE)
  
  p <- payoff_matrix(N_opts)
  idx <- which(p>0.5,arr.ind = TRUE,useNames = FALSE)
  rps_wins <- list()
  for (i in c(1:nrow(idx))){
    rps_wins[[i]] <- c(idx[i,1],idx[i,2])
  }
  
  h <- sample(rps_wins,1)[[1]]
  
  ## Luckier player to win
  dt <- thetas[t_ids[1]] - thetas[t_ids[2]]
  w <- rbinom(1,1,plogis(dt))
  if(w==0){
    h <- c(h[2],h[1])
  }
  res <- c(t_ids[1],t_ids[2],h[1],h[2],w,w)
  return(res)
}

simulate_NT_model <- function(thetas,phis,N_opts,replace=FALSE){
  t_ids <- sample(c(1:length(thetas)),2,replace=FALSE)
  phi_ids <- sample(c(1:length(phis)),2,replace=FALSE)
  h <- sample(c(1:N_opts),2,replace=replace)
  
  d_theta <- thetas[t_ids[1]] - thetas[t_ids[2]]
  d_phi <- phis[h[1]] - phis[h[2]]
  d_phi2 <- atan2(sinpi(d_phi),cospi(d_phi)) %% (2*pi)
  g_d_phi <- 2*sin(d_phi2)
  # g_d_phi <- 2*atanh(sinpi(d_phi))
  
  logit <- d_theta + g_d_phi 
  
  p <- plogis(logit)
  w <- rbinom(1,1,p)
  
  res <- c(t_ids[1],t_ids[2],h[1],h[2],w,p)
  return(res)
}

simulate_dice_rolls <- function(thetas,dice,replace=FALSE){
  t_ids <- sample(c(1:length(thetas)),2,replace=FALSE)
  N_opts <- length(dice)
  
  h <- sample(c(1:N_opts),2,replace=replace)
  roll_1 <- sample(dice[[h[1]]],1)
  roll_1_w_handicap <- roll_1 + thetas[t_ids[1]]
  roll_2 <- sample(dice[[h[2]]],1)
  roll_2_w_handicap <- roll_2 + thetas[t_ids[2]]
  
  p <- mean(outer(dice[[h[1]]]+ thetas[t_ids[1]],dice[[h[2]]]+ thetas[t_ids[2]],">="))
  
  w <- ifelse(roll_1_w_handicap >= roll_2_w_handicap,1,0)
  res <- c(t_ids[1],t_ids[2],h[1],h[2],w,p)
  return(res)
}

```

## Estimation Functions

```{r}


LogLikELO <- function(data, theta) {
  p <- plogis(theta[data$player1] - theta[data$player2])
  ll <- sum(data$result * log(p) + (1 - data$result) * log(1 - p))
  return(ll)
}

fit_elo_model <- function(N_players,matches){
  
  starting_theta <- rnorm(N_players)
  
  sol <- optim(
    starting_theta,
    LogLikELO,
    data = matches,
    method = 'L-BFGS-B',
    lower = -3,
    upper = 3,
    control = list(fnscale = -1, maxit = 1e6)
  )
  
  return(sol)
}


LogLik_NT <- function(params, data, N_ids,
                      theta_centering = TRUE,
                      ridge_penalty = TRUE) {
  
    ## Mean centering or setting first theta to 0
    if (theta_centering) {
        theta <- params[1:N_ids]
        theta <- theta - mean(theta)
        pos <- params[(N_ids + 1):length(params)]
        n_pos <- length(params) - N_ids
    } else {
        theta <- c(0, params[1:(N_ids - 1)])
        pos <- params[N_ids:length(params)]
        n_pos <- length(params) - (N_ids + 1)
    }
    
    d_theta <- theta[data$player1] - theta[data$player2]

    N_Pmat <- (1+sqrt(1+8*n_pos))/2
    P <- matrix(0, nrow = N_Pmat, ncol = N_Pmat)
    P[upper.tri(P)] <- pos
    P <- P - t(P)
      
    d_phi <- P[data$hand1,data$hand2]
    logit <- d_theta + d_phi
    
    p <- plogis(logit)
    ll <- sum(data$result * log(p) + (1 - data$result) * log(1 - p))
    
    penalty <- ifelse(ridge_penalty, 0.1 * sum(theta^2),0)
    neg_ll <- -ll + penalty
    
    if (is.finite(neg_ll)) neg_ll else 1e10
}

estimate_nt_elo <- function(N_ids, N_opts, matches,
                            theta_centering = TRUE,
                            ridge_penalty = TRUE) {
    ## Initializing params
    n_theta <- ifelse(theta_centering,N_ids,N_ids-1)
    starting_theta <- rnorm(n_theta)

    n_payoff <- N_opts*(N_opts-1)/2
    starting_po <- rnorm(n_payoff)
    starting_params <- c(starting_theta, starting_po)

    ## Solver
    sol <- optim(
      starting_params,
      LogLik_NT,
      data = matches,
      N_ids = N_ids,
      theta_centering = theta_centering,
      ridge_penalty = ridge_penalty,
      method = 'L-BFGS-B',
      hessian = TRUE,
      lower = rep(-Inf,(n_theta+n_payoff)),
      upper = rep(Inf,(n_theta+n_payoff)),
      control = list(fnscale = 1, maxit = 1e6)
    )
    return(sol)
  }

get_payoff_mat <- function(x){
    n <- (1+sqrt(1+8*length(x)))/2
    P <- matrix(0,n,n)
    P[upper.tri(P)] <- x
    P <- P - t(P)
    return(P)
}
```

## RPS Simulation

### Simple case: deterministic outcomes

```{r}
set.seed(1729)
N_ids <- 100
N_games <- 1000
N_opts <- 3 ## For rock-paper-scissors
payoff <- payoff_matrix(N_opts,eps=0)
rps_thetas <- rep(0,N_ids)

rps_matches <- as.data.frame(do.call(rbind, replicate(N_games,simulate_rps(rps_thetas,N_opts,payoff),simplify = FALSE)))
rps_matches <- rps_matches |> rename(player1="V1",
                             player2="V2",
                             hand1 = "V3",
                             hand2 = "V4",
                             result="V5",
                             true_p="V6")

```

```{r}
# est <- estimate_nt_elo(N_ids,N_opts,rps_matches)
print(plogis(get_payoff_mat(tail(est$par,3))))
```

```{r}
a <- 2
a

```