---
title: "Measurement Project"
format: pdf
editor: visual
---

## Introduction

ELO models are widely used, but have a key underlying assumption that the difference in latent skill/ability is the only thing that contributes to probability of winning. This inherently means that probabilities are transitive, i.e. P(i \> j) and P(j \> k) =\> P(i \>k).

This works for a lot of games, such as chess, where the ELO model is widely used. Improvements to ELO are also popular taking into account Rating Deviation, the uncertainty of an individual's ability, estimated based on the consistency of results and the number of games played.

```{r}
library(tidyverse)
library(future.apply)
library(dplyr)
library(ggplot2)
library(patchwork)
library(latex2exp)
library(see)
options(warn = -1)
```

### Game simulation functions

```{r}

simulate_elo_game <- function(thetas,N_opts=3){
  t_ids <- sample(c(1:length(thetas)),2,replace=FALSE)
  t <- thetas[t_ids]
  dt <- t[1]-t[2]
  p <- 1/(1+exp(-dt))
  w <- rbinom(n=1,size=1,prob=p)
  ## Hand played is randomly chosen
  h <- sample(c(1:N_opts),2,replace=TRUE)
  res <- c(t_ids[1],t_ids[2],h[1],h[2],w,p)
  return(res)
}

## Non-transitive payoff probability matrix (eps -> [0,2])
payoff_matrix <- function(N,eps=0) {
  eps <- abs(eps)
  # generate square matrix with 0s
  
  A <- matrix(0, nrow = N, ncol = N)
  
  # Fill upper triangle with payoff
  for(i in c(1:(N-1))){
    A[i,(i+1):N] <- rep(c(-1+eps, 1-eps), length.out = (N-i))
  }
  
  # Enforce skew-symmetry for cyclic game
  A <- A - t(A)
  
  # Convert to Probabilities
  pA <- 0.5*(A + 1)
  return(pA)
}


simulate_rps <- function(thetas,N_opts,payoff=NULL,replace=FALSE){
  t_ids <- sample(c(1:length(thetas)),2,replace=FALSE)
  if(is.null(payoff)){
    p <- payoff_matrix(N_opts,eps=0)
  }else{
    p <- payoff
  }
  
  h <- sample(c(1:N_opts),2,replace=replace)
  w <- rbinom(1,1,p[h[1],h[2]])
  # w <- ifelse(prob>0,rbinom(1,1,prob),rbinom(1,1,1+prob))
  res <- c(t_ids[1],t_ids[2],h[1],h[2],w,p[h[1],h[2]])
  return(res)
}

simulate_rps_luck <- function(thetas,N_opts){
  t_ids <- sample(c(1:length(thetas)),2,replace=FALSE)
  
  p <- payoff_matrix(N_opts)
  idx <- which(p>0.5,arr.ind = TRUE,useNames = FALSE)
  rps_wins <- list()
  for (i in c(1:nrow(idx))){
    rps_wins[[i]] <- c(idx[i,1],idx[i,2])
  }
  
  h <- sample(rps_wins,1)[[1]]
  
  ## Luckier player to win
  dt <- thetas[t_ids[1]] - thetas[t_ids[2]]
  w <- rbinom(1,1,plogis(dt))
  if(w==0){
    h <- c(h[2],h[1])
  }
  res <- c(t_ids[1],t_ids[2],h[1],h[2],w,w)
  return(res)
}

simulate_NT_model <- function(thetas,phis,N_opts,replace=FALSE){
  t_ids <- sample(c(1:length(thetas)),2,replace=FALSE)
  phi_ids <- sample(c(1:length(phis)),2,replace=FALSE)
  h <- sample(c(1:N_opts),2,replace=replace)
  
  d_theta <- thetas[t_ids[1]] - thetas[t_ids[2]]
  d_phi <- phis[h[1]] - phis[h[2]]
  d_phi2 <- atan2(sinpi(d_phi),cospi(d_phi)) %% (2*pi)
  g_d_phi <- 2*sin(d_phi2)
  # g_d_phi <- 2*atanh(sinpi(d_phi))
  
  logit <- d_theta + g_d_phi 
  
  p <- plogis(logit)
  w <- rbinom(1,1,p)
  
  res <- c(t_ids[1],t_ids[2],h[1],h[2],w,p)
  return(res)
}

simulate_dice_rolls <- function(thetas,dice,replace=FALSE){
  t_ids <- sample(c(1:length(thetas)),2,replace=FALSE)
  N_opts <- length(dice)
  
  h <- sample(c(1:N_opts),2,replace=replace)
  roll_1 <- sample(dice[[h[1]]],1)
  roll_1_w_handicap <- roll_1 + thetas[t_ids[1]]
  roll_2 <- sample(dice[[h[2]]],1)
  roll_2_w_handicap <- roll_2 + thetas[t_ids[2]]
  
  p <- mean(outer(dice[[h[1]]]+ thetas[t_ids[1]],dice[[h[2]]]+ thetas[t_ids[2]],">="))
  
  w <- ifelse(roll_1_w_handicap >= roll_2_w_handicap,1,0)
  res <- c(t_ids[1],t_ids[2],h[1],h[2],w,p)
  return(res)
}

```

### Estimation functions

```{r}
LogLikELO <- function(data, theta) {
  p <- plogis(theta[data$player1] - theta[data$player2])
  ll <- sum(data$result * log(p) + (1 - data$result) * log(1 - p))
  return(ll)
}

fit_elo_model <- function(N_players,matches){
  
  starting_theta <- rnorm(N_players)
  
  sol <- optim(
    starting_theta,
    LogLikELO,
    data = matches,
    method = 'L-BFGS-B',
    lower = -3,
    upper = 3,
    control = list(fnscale = -1, maxit = 1e6)
  )
  
  return(sol)
}


LogLik_NT <- function(params, data, N_ids,
                      theta_centering = TRUE,
                      use_atanh = FALSE,
                      scale_by_sd = FALSE,
                      ridge_penalty = TRUE) {
  
  ## Mean centering or setting first theta to 0
  if (theta_centering) {
    theta <- params[1:N_ids]
    theta <- theta - mean(theta)
    phi <- c(0, params[(N_ids + 1):length(params)])
  } else {
    theta <- c(0, params[1:(N_ids - 1)])
    phi <- c(0, params[N_ids:length(params)])
  }

  sigma_theta <- sd(theta)
  d_theta <- theta[data$player1] - theta[data$player2]

  d_phi <- phi[data$hand1] - phi[data$hand2]
  d_phi <- atan2(sinpi(d_phi), cospi(d_phi)) %% (2*pi)

  if (use_atanh) {
    g_d_phi <- 2 * atanh(sin(d_phi))
  } else {
    g_d_phi <- 2 * sin(d_phi)
  }

  if (scale_by_sd) {
    g_d_phi <- g_d_phi * sigma_theta
  }

  logit <- d_theta + g_d_phi
  p <- plogis(logit)
  ll <- sum(data$result * log(p) + (1 - data$result) * log(1 - p))

  penalty <- ifelse(ridge_penalty, 0.1 * sum(theta^2),0)
  neg_ll <- -ll + penalty

  if (is.finite(neg_ll)) neg_ll else 1e10
}

estimate_nt_elo <- function(N_ids, N_opts, matches, n_start = 5,
                            theta_centering = "mean",
                            use_atanh = FALSE,
                            scale_by_sd = FALSE,
                            ridge_penalty = TRUE) {
  n_theta <- ifelse(theta_centering,N_ids,N_ids-1)

  opt_sol <- NA
  opt_sol_val <- Inf
  opt_start <- NA

  for (k in c(1:(n_start + 2))) {
    starting_theta <- rnorm(n_theta)

    if (k == 1) {
      starting_phi <- rep(0.25, N_opts - 1)
    } else if (k == 2) {
      starting_phi <- c(1:(N_opts - 1)) * 2 / N_opts
      starting_phi <- runif(N_opts - 1, min = 0, max = 2)
    } else {
      starting_phi <- runif(N_opts - 1, min = 0, max = 2)
    }

    starting_params <- c(starting_theta, starting_phi)
    sol <- optim(
      starting_params,
      LogLik_NT,
      data = matches,
      N_ids = N_ids,
      theta_centering = theta_centering,
      use_atanh = use_atanh,
      scale_by_sd = scale_by_sd,
      ridge_penalty = ridge_penalty,
      method = 'L-BFGS-B',
      hessian = TRUE,
      lower = c(rep(-Inf, n_theta), rep(0, N_opts - 1)),
      upper = c(rep(Inf, n_theta), rep(2, N_opts - 1)),
      control = list(fnscale = 1, maxit = 1e6)
    )
    if (sol$value < opt_sol_val) {
      opt_sol_val <- sol$value
      opt_sol <- sol
      opt_start <- starting_phi
    }
  }

  return(list(sol = opt_sol, starting_phi = opt_start))
}
```

### Plotting functions

```{r}
plot_elo <- function(sol,thetas,p_true, p_pred,w){
  df <- data.frame(theta=thetas,theta_est=sol$par)
  p_elo <- ggplot(df,aes(x=theta,y=theta_est)) + 
    geom_point(color=okabeito_colors(3)) + 
    theme_bw() + 
    labs(x=TeX("$\\theta$"),y=TeX("$\\hat{\\theta}$")) + 
    annotate("text", x = 2, y = -3, 
             label = paste("r =", round(cor(df$theta,df$theta_est,method="spearman"),2)),
             size=4) + ggtitle("True vs. Estimated thetas (ELO Model)")
  
  
  if (!is.null(p_pred) && !is.null(p_true)) {
    p_scatter <- ggplot(data.frame(p_pred=p_pred, p_true=p_true),
                        aes(x=p_pred, y=p_true,color=as.factor(w))) +
      geom_point(size=1, alpha=0.5) +
      theme_bw() +
      labs(x="Predicted Prob", y="True p / Result", title="Predicted vs True",colour="Result")
    return(p_elo | p_scatter)
  }
  return(p_elo)
}

plot_nt_elo <- function(sol, thetas, N_ids, N_opts, theta_centering=TRUE,
                        labels=c("Rock", "Paper", "Scissors"),
                        p_pred=NULL, p_true=NULL,w=NULL){
  
  
  if (theta_centering){
    est_th <- sol$par[1:N_ids]
    est_th <- est_th - mean(est_th)
    theta_offset <- 0
  } else {
    est_th <- c(0, sol$par[1:(N_ids - 1)])
    theta_offset <- 1
  }

  est_theta <- as.data.frame(list(est=est_th, true=thetas))

  est_phi <- as.data.frame(list(
    est = c(0, sol$par[(N_ids + 1 - theta_offset):length(sol$par)]),
    r   = rep(1, N_opts),
    id  = c(1:N_opts)
  ))
  est_phi <- est_phi |> mutate(phi = pi * (est %% 2))

  est_theta_mu  <- round(mean(est_theta$est), 2)
  est_theta_var <- round(sd(est_theta$est)^2, 2)

  p_phi <- ggplot(est_phi) +
    geom_point(aes(r, phi, color=factor(id)), size=2) +
    coord_radial(theta="y",
                 start=pi/2,
                 rlim=c(0,1),
                 thetalim=c(0, 2*pi),
                 reverse = "theta",
                 expand=FALSE) +
    scale_y_continuous(breaks=c(0,pi/2,pi,3*pi/2,2*pi),
                       labels = c("0", "π/2", "π", "3π/2", "2π")) +
    ggtitle("Phi") +
    labs(color="Hand",x="",y="") +
    scale_color_discrete(labels=labels)

  scatter_title <- paste0("Theta~N(", est_theta_mu, ",", est_theta_var, "), r=",
         round(cor(est_theta$est, est_theta$true, method="spearman"), 2))
  
  p_theta <- ggplot(est_theta, aes(x=true, y=est)) +
    geom_point() +
    theme_bw() +
    ggtitle(str_wrap(scatter_title,20))

  if (!is.null(p_pred) && !is.null(p_true)) {
    p_scatter <- ggplot(data.frame(p_pred=p_pred, p_true=p_true),
                        aes(x=p_pred, y=p_true,color=as.factor(w))) +
      geom_point(size=1, alpha=0.5) +
      theme_bw() +
      labs(x="Predicted prob", y="True prob", title="Predicted vs True",colour="Result")
    return(p_theta | p_phi | p_scatter)
  }

  return(p_theta | p_phi)
}

get_predicted_prob <- function(matches,theta,phi=NULL,use_atanh=FALSE,scale_by_sd=FALSE,use_theta=TRUE){
  if(is.null(phi)){
    phi <- rep(0,max(c(matches$hand1,matches$hand2)))
  }
  
  d_theta <- theta[matches$player1] - theta[matches$player2]
  d_phi <- phi[matches$hand1] - phi[matches$hand2]
  
  d_phi2 <- atan2(sinpi(d_phi),cospi(d_phi)) %% (2*pi)
  
  if(use_atanh){
    g_d_phi <- 2*atanh(sin(d_phi2))
  }else{
    g_d_phi <- 2*sin(d_phi2)
  }
  
  logit <- use_theta*d_theta + g_d_phi 
  pred_p <- plogis(logit)
  return(pred_p)
}

get_res_estimates <- function(res,N_ids,theta_centering){
  if(theta_centering){
    res_theta <- res$sol$par[1:N_ids]
    res_theta <- res_theta - mean(res_theta)
    res_phi <- c(0,res$sol$par[(N_ids+1):length(res$sol$par)])
  }else{
    res_theta <- c(0,res$sol$par[1:(N_ids-1)])
    res_phi <- c(0,res$sol$par[N_ids:length(res$sol$par)])
  }
  return(list(theta=res_theta,phi=res_phi))
}

```

# Simulating Games and Testing model fits

## ELO Data

```{r}
## Simulating elo game data
N_ids <- 100
N_games <- 1000
elo_thetas <- rnorm(N_ids)

elo_matches <- as.data.frame(do.call(rbind, replicate(N_games,simulate_elo_game(elo_thetas),simplify = FALSE)))

elo_matches <- elo_matches |> rename(player1="V1",
                             player2="V2",
                             hand1 = "V3",
                             hand2 = "V4",
                             result="V5",
                             true_p="V6")
```

```{r}
elo_sol <- fit_elo_model(N_ids,elo_matches)
elo_pred <- get_predicted_prob(elo_matches,theta=elo_sol$par)
plot_elo(elo_sol,elo_thetas,p_true=elo_matches$true_p,p_pred=elo_pred,w=elo_matches$result)

res_nt <- estimate_nt_elo(N_ids,N_opts=3,elo_matches,n_start = 10,
                          theta_centering = TRUE,
                          use_atanh = FALSE,
                          scale_by_sd = FALSE,
                          ridge_penalty = TRUE)

estimates <- get_res_estimates(res_nt,N_ids,theta_centering=TRUE)
pred_p <- get_predicted_prob(elo_matches,theta=estimates$theta,phi=estimates$phi)

plot_nt_elo(res_nt$sol,elo_thetas,N_ids,N_opts=3,theta_centering = TRUE,
            p_true=elo_matches$true_p,p_pred=pred_p,w=elo_matches$result)

res_nt <- estimate_nt_elo(N_ids,N_opts=3,elo_matches,n_start = 10,
                          theta_centering=FALSE,use_atanh=TRUE)

estimates <- get_res_estimates(res_nt,N_ids,theta_centering=FALSE)
pred_p <- get_predicted_prob(elo_matches,
                             theta=estimates$theta,phi=estimates$phi,
                             use_atanh = TRUE)

plot_nt_elo(res_nt$sol,elo_thetas,N_ids,N_opts=3,theta_centering = FALSE,
            p_true=elo_matches$true_p,p_pred=pred_p,w=elo_matches$result)


```

Weird outliers in theta estimation make me believe that some ridge penalty on theta would be good to have. Getting Reasonable estimates from the non-transitive models, d_phis are small (or close to pi), but not exactly 0 or pi (what I would expect)

## Rock-Paper-Scissors

```{r}
## Simulating the game data
set.seed(1729)
N_ids <- 100
N_games <- 1000
N_opts <- 7 ## For rock-paper-scissors
payoff_matrix <- payoff_matrix
rps_thetas <- rep(0,N_ids)

rps_matches <- as.data.frame(do.call(rbind, replicate(N_games,simulate_rps(rps_thetas,N_opts),simplify = FALSE)))

rps_matches <- rps_matches |> rename(player1="V1",
                             player2="V2",
                             hand1 = "V3",
                             hand2 = "V4",
                             result="V5",
                             true_p="V6")
```

```{r}

elo_sol <- fit_elo_model(N_ids,rps_matches)
elo_pred <- get_predicted_prob(rps_matches,theta=elo_sol$par)
plot_elo(elo_sol,rps_thetas,p_true=rps_matches$true_p,p_pred=elo_pred,w=rps_matches$result)

res_nt <- estimate_nt_elo(N_ids,N_opts,rps_matches,n_start = 10,
                          theta_centering = TRUE,
                          use_atanh = FALSE,
                          scale_by_sd = FALSE,
                          ridge_penalty = TRUE)

estimates <- get_res_estimates(res_nt,N_ids,theta_centering=TRUE)
pred_p <- get_predicted_prob(rps_matches,theta=estimates$theta,phi=estimates$phi)


plot_nt_elo(res_nt$sol,rps_thetas,N_ids,N_opts,theta_centering = TRUE,
            p_true=rps_matches$true_p,p_pred=pred_p,w=rps_matches$result,
            labels=c("Rock", "Paper", "Scissors","Spock","Lizard"))

res_nt <- estimate_nt_elo(N_ids,N_opts,rps_matches,n_start = 10,
                          theta_centering=FALSE,use_atanh=TRUE)


estimates <- get_res_estimates(res_nt,N_ids,theta_centering=FALSE)
pred_p <- get_predicted_prob(rps_matches,
                             theta=estimates$theta,phi=estimates$phi,
                             use_atanh = TRUE)

plot_nt_elo(res_nt$sol,rps_thetas,N_ids,N_opts,theta_centering = FALSE,
            p_true=rps_matches$true_p,p_pred=pred_p,w=rps_matches$result,
            labels=c("Rock", "Paper", "Scissors","Spock","Lizard"))


# ## Fitting ELO model to RPS Data
# rps_elo_sol <- fit_elo_model(N_ids,rps_matches)
# plot_elo(rps_elo_sol,rps_thetas)
# ## Fitting simple non-transitive model
# res_nt <- estimate_nt_elo(N_ids, N_opts, rps_matches)
# plot_nt_elo(res_nt$sol,rps_thetas,N_ids,N_opts,labels=c("Rock", "Paper", "Scissors","Spock","Lizard"))
# ## Fitting experiemental non-transitive model
# res_nt <- estimate_nt_elo(N_ids, N_opts, rps_matches,theta_centering="fixed",use_atanh=TRUE)
# plot_nt_elo(res_nt$sol,rps_thetas,N_ids,N_opts,theta_offset=1,labels=c("Rock", "Paper", "Scissors","Spock","Lizard"))

```

```{r}
## Simulating the game data
set.seed(1729)
N_ids <- 100
N_games <- 1000
N_opts <- 5 ## For rock-paper-scissors
rps_thetas <- rep(0,N_ids)

rps_matches <- as.data.frame(do.call(rbind, replicate(N_games,simulate_rps(rps_thetas,N_opts),simplify = FALSE)))

rps_matches <- rps_matches |> rename(player1="V1",
                             player2="V2",
                             hand1 = "V3",
                             hand2 = "V4",
                             result="V5",
                             true_p="V6")
```

```{r}
elo_sol <- fit_elo_model(N_ids,rps_matches)
elo_pred <- get_predicted_prob(rps_matches,theta=elo_sol$par)
plot_elo(elo_sol,rps_thetas,p_true=rps_matches$true_p,p_pred=elo_pred,w=rps_matches$result)

res_nt <- estimate_nt_elo(N_ids,N_opts,rps_matches,n_start = 10,
                          theta_centering = TRUE,
                          use_atanh = FALSE,
                          scale_by_sd = FALSE,
                          ridge_penalty = TRUE)

estimates <- get_res_estimates(res_nt,N_ids,theta_centering=TRUE)
pred_p <- get_predicted_prob(rps_matches,theta=estimates$theta,phi=estimates$phi)

plot_nt_elo(res_nt$sol,rps_thetas,N_ids,N_opts,theta_centering = TRUE,
            p_true=rps_matches$true_p,p_pred=pred_p,w=rps_matches$result,
            labels=c("Rock", "Paper", "Scissors","Spock","Lizard"))

res_nt <- estimate_nt_elo(N_ids,N_opts,rps_matches,n_start = 10,
                          theta_centering=FALSE,use_atanh=TRUE)

estimates <- get_res_estimates(res_nt,N_ids,theta_centering=FALSE)
pred_p <- get_predicted_prob(rps_matches,
                             theta=estimates$theta,phi=estimates$phi,
                             use_atanh = TRUE)

plot_nt_elo(res_nt$sol,rps_thetas,N_ids,N_opts,theta_centering = FALSE,
            p_true=rps_matches$true_p,p_pred=pred_p,w=rps_matches$result,
            labels=c("Rock", "Paper", "Scissors","Spock","Lizard"))

# ## Fitting ELO model to RPS Data
# rps_elo_sol <- fit_elo_model(N_ids,rps_matches)
# plot_elo(rps_elo_sol,rps_thetas)
# ## Fitting simple non-transitive model
# res_nt <- estimate_nt_elo(N_ids, N_opts, rps_matches)
# plot_nt_elo(res_nt$sol,rps_thetas,N_ids,N_opts,labels=c("Rock", "Paper", "Scissors","Spock","Lizard"))
# ## Fitting experiemental non-transitive model
# res_nt <- estimate_nt_elo(N_ids, N_opts, rps_matches,theta_centering="fixed",use_atanh=TRUE)
# plot_nt_elo(res_nt$sol,rps_thetas,N_ids,N_opts,theta_offset=1,labels=c("Rock", "Paper", "Scissors","Spock","Lizard"))
```

### RPS with "luck"

```{r}
N_ids <- 100
N_opts <- 5
N_games <- 1000
rps_luck_thetas <- rnorm(N_ids,mean=0,sd=1)
rps_luck_matches <- as.data.frame(do.call(rbind, replicate(N_games,simulate_rps_luck(rps_luck_thetas,N_opts),simplify = FALSE)))

rps_luck_matches <- rps_luck_matches |> rename(player1="V1",
                             player2="V2",
                             hand1 = "V3",
                             hand2 = "V4",
                             result="V5",
                             true_p="V6")
```

```{r}
elo_sol <- fit_elo_model(N_ids,rps_luck_matches)
elo_pred <- get_predicted_prob(rps_luck_matches,theta=elo_sol$par)
plot_elo(elo_sol,rps_luck_thetas,p_true=rps_luck_matches$true_p,p_pred=elo_pred,w=rps_luck_matches$result)

res_nt <- estimate_nt_elo(N_ids,N_opts,rps_luck_matches,n_start = 10,
                          theta_centering = TRUE,
                          use_atanh = FALSE,
                          scale_by_sd = FALSE,
                          ridge_penalty = TRUE)

estimates <- get_res_estimates(res_nt,N_ids,theta_centering=TRUE)
pred_p <- get_predicted_prob(rps_luck_matches,theta=estimates$theta,phi=estimates$phi)

plot_nt_elo(res_nt$sol,rps_luck_thetas,N_ids,N_opts,theta_centering = TRUE,
            p_true=rps_luck_matches$true_p,p_pred=pred_p,w=rps_luck_matches$result,
            labels=c("Rock", "Paper", "Scissors","Spock","Lizard"))

res_nt <- estimate_nt_elo(N_ids,N_opts,rps_luck_matches,n_start = 10,
                          theta_centering=FALSE,use_atanh=TRUE)

estimates <- get_res_estimates(res_nt,N_ids,theta_centering=FALSE)
pred_p <- get_predicted_prob(rps_luck_matches,
                             theta=estimates$theta,phi=estimates$phi,
                             use_atanh = TRUE)

plot_nt_elo(res_nt$sol,rps_luck_thetas,N_ids,N_opts,theta_centering = FALSE,
            p_true=rps_luck_matches$true_p,p_pred=pred_p,w=rps_luck_matches$result,
            labels=c("Rock", "Paper", "Scissors","Spock","Lizard"))

# res_elo_rps_luck <- fit_elo_model(N_ids,rps_luck_matches)
# res_rps_luck     <- estimate_nt_elo(N_ids, N_opts, rps_luck_matches, 10)
# res_rps_luck_exp <- estimate_nt_elo(N_ids, N_opts, rps_luck_matches, 10, theta_centering="fixed", use_atanh=TRUE)
# 
# est_thetas <- as_tibble(res_elo_rps_luck$par) |> rename(elo_theta='value') |> mutate(id=row_number())
# est_thetas$rps_theta     <- res_rps_luck$sol$par[1:N_ids]
# est_thetas$rps_exp_theta <- c(0, res_rps_luck_exp$sol$par[1:(N_ids-1)])
# 
# est_phis <- as_tibble(c(1:N_opts)) |> rename(hand='value')
# est_phis$rps_phi     <- c(0, res_rps_luck$sol$par[(N_ids+1):(N_ids+N_opts-1)])
# est_phis$rps_exp_phi <- c(0, res_rps_luck_exp$sol$par[N_ids:(N_ids+N_opts-2)])
# 
# rps_luck_matches_predict <- rps_luck_matches |>
#   left_join(est_thetas, by=join_by(player1==id)) |>
#   left_join(est_thetas, by=join_by(player2==id)) |>
#   left_join(est_phis,   by=join_by(hand1==hand)) |>
#   left_join(est_phis,   by=join_by(hand2==hand)) |>
#   mutate(elo_d_theta     = elo_theta.x - elo_theta.y,
#          rps_d_theta     = rps_theta.x - rps_theta.y,
#          rps_exp_d_theta = rps_exp_theta.x - rps_exp_theta.y,
#          rps_d_phi       = rps_phi.x - rps_phi.y,
#          rps_exp_d_phi   = rps_exp_phi.x - rps_exp_phi.y,
#          rps_g_d_phi     = 2*sin(atan2(sinpi(rps_d_phi), cospi(rps_d_phi))),
#          rps_exp_g_d_phi = 2*atanh(sin(atan2(sinpi(rps_exp_d_phi), cospi(rps_exp_d_phi)))),
#          p_elo     = plogis(elo_d_theta),
#          p_rps     = plogis(rps_d_theta + rps_g_d_phi),
#          p_rps_exp = plogis(rps_exp_d_theta + rps_exp_g_d_phi))
# 
# plot_elo(res_elo_rps_luck, rps_luck_thetas)
# plot_nt_elo(res_rps_luck$sol, rps_luck_thetas, N_ids, N_opts,
#             labels=c("Rock","Paper","Scissors","Spock","Lizard"),
#             p_pred=rps_luck_matches_predict$p_rps,
#             p_true=rps_luck_matches_predict$result)
# plot_nt_elo(res_rps_luck_exp$sol, rps_luck_thetas, N_ids, N_opts, theta_offset=1,
#             labels=c("Rock","Paper","Scissors","Spock","Lizard"),
#             p_pred=rps_luck_matches_predict$p_rps_exp,
#             p_true=rps_luck_matches_predict$result)
```

## Non-Transitive Dice game

```{r}
grime_3 <- c(7/12,11/36,7/12) #Red, Blue, Olive
grime_4 <- c(2/3,1/3,4/9,1/2,2/3,1/3) #Red, Blue, Olive, Yellow
grime_5 <- c(7/12,11/36,13/18,4/9,7/12,1/3,2/3,5/9,5/18,5/9) #Red, Blue, Olive, Yellow, Magenta


grime_mats_list <- list(grime_3,grime_4,grime_5)

grime_dice_payoffs <- function(k,grime_mats=grime_mats_list){
  p <- matrix(0,k,k)
  upper_tri_idx <- as.data.frame(which(upper.tri(p),arr.ind = TRUE)) |> 
    arrange(row,col) |> 
    as.matrix()
  
  lower_tri_idx <- as.data.frame(which(lower.tri(p),arr.ind = TRUE)) |> 
    arrange(col,row) |> 
    as.matrix()
  
  p[upper_tri_idx] <- grime_mats[[k-2]]
  p[lower_tri_idx] <- 1-p[upper_tri_idx]
  diag(p) <- 1/2  
  return(p)
}

```

### 3-dice

```{r}
N_games <- 1000
N_ids <- 100
N_opts <- 3
thetas <- rnorm(N_ids)

payoff_mat <- grime_dice_payoffs(N_opts)
grime_dice_matches <- as.data.frame(
  do.call(rbind, 
          replicate(N_games,
                    simulate_rps(thetas,N_opts,payoff_mat,TRUE),
                    simplify = FALSE)))

grime_dice_matches <- grime_dice_matches |> rename(player1="V1",
                             player2="V2",
                             hand1 = "V3",
                             hand2 = "V4",
                             result="V5",
                             true_p="V6")
```

```{r}


elo_sol <- fit_elo_model(N_ids,grime_dice_matches)
elo_pred <- get_predicted_prob(grime_dice_matches,theta=elo_sol$par)
plot_elo(elo_sol,thetas,p_true=grime_dice_matches$true_p,p_pred=elo_pred,w=grime_dice_matches$result)

res_nt <- estimate_nt_elo(N_ids,N_opts,grime_dice_matches,n_start = 10,
                          theta_centering = TRUE,
                          use_atanh = FALSE,
                          scale_by_sd = FALSE,
                          ridge_penalty = TRUE)

estimates <- get_res_estimates(res_nt,N_ids,theta_centering=TRUE)
pred_p <- get_predicted_prob(grime_dice_matches,theta=estimates$theta,phi=estimates$phi)

plot_nt_elo(res_nt$sol,thetas,N_ids,N_opts,theta_centering = TRUE,
            p_true=grime_dice_matches$true_p,p_pred=pred_p,w=grime_dice_matches$result,
            labels=c("Red", "Blue", "Olive", "Yellow", "Magenta"))

est_thetas <- estimates$theta
est_phis <- estimates$phi
post_estimation <- grime_dice_matches |>
  mutate(d_theta = est_thetas[player1]-est_thetas[player2],
         d_phi = est_phis[hand1]-est_phis[hand2],
         d_phi2 = atan2(sinpi(d_phi), cospi(d_phi)) %% (2*pi),
         exponent = -(2*sin(d_phi2)),
         prob = 1/(1+exp(exponent))) |>
  group_by(hand1,hand2) |>
  summarise(expected_payoff=mean(prob),.groups = "drop_last")

payoff_mat
est_payoff_mat <- xtabs(expected_payoff ~ hand1 + hand2, data=post_estimation)
est_payoff_mat


res_nt <- estimate_nt_elo(N_ids,N_opts,grime_dice_matches,n_start = 10,
                          theta_centering=FALSE,use_atanh=TRUE)

estimates <- get_res_estimates(res_nt,N_ids,theta_centering=FALSE)
pred_p <- get_predicted_prob(grime_dice_matches,
                             theta=estimates$theta,phi=estimates$phi,
                             use_atanh = TRUE)

plot_nt_elo(res_nt$sol,thetas,N_ids,N_opts,theta_centering = FALSE,
            p_true=grime_dice_matches$true_p,p_pred=pred_p,w=grime_dice_matches$result,
            labels=c("Red", "Blue", "Olive", "Yellow", "Magenta"))

est_thetas <- estimates$theta
est_phis <- estimates$phi
post_estimation <- grime_dice_matches |>
  mutate(d_theta = est_thetas[player1]-est_thetas[player2],
         d_phi = est_phis[hand1]-est_phis[hand2],
         d_phi2 = atan2(sinpi(d_phi), cospi(d_phi)) %% (2*pi),
         exponent = -(2*atanh(sin(d_phi2))),
         prob = 1/(1+exp(exponent))) |>
  group_by(hand1,hand2) |>
  summarise(expected_payoff=mean(prob),.groups = "drop_last")

payoff_mat
est_payoff_mat <- xtabs(expected_payoff ~ hand1 + hand2, data=post_estimation)
est_payoff_mat

# res <- estimate_nt_elo(N_ids,N_opts,grime_dice_matches,n_start=10)
# plot_nt_elo(res$sol,thetas,N_ids,N_opts,labels=c("Red", "Blue", "Olive", "Yellow", "Magenta"))
# 
# est_thetas <- c(0,res$sol$par[1:(N_ids-1)])
# est_phis <- c(0,tail(res$sol$par,N_opts-1))
# post_estimation <- grime_dice_matches |> 
#   mutate(d_theta = est_thetas[player1]-est_thetas[player2],
#          d_phi = est_phis[hand1]-est_phis[hand2],
#          d_phi2 = atan2(sin(d_phi),cos(d_phi)),
#          exponent = -(2*sinpi(d_phi2)),
#          prob = 1/(1+exp(exponent))) |>
#   group_by(hand1,hand2) |>
#   summarise(expected_payoff=mean(prob),.groups = "drop_last")
# 
# payoff_mat
# est_payoff_mat <- xtabs(expected_payoff ~ hand1 + hand2, data=post_estimation)
# est_payoff_mat


# res <- estimate_nt_elo(N_ids,N_opts,grime_dice_matches,n_start=10,theta_centering="fixed",use_atanh=TRUE)
# plot_nt_elo(res$sol,thetas,N_ids,N_opts,labels=c("Red", "Blue", "Olive", "Yellow", "Magenta"))
# 
# est_thetas <- c(0,res$sol$par[1:(N_ids-1)])
# est_phis <- c(0,tail(res$sol$par,N_opts-1))
# post_estimation <- grime_dice_matches |> 
#   mutate(d_theta = est_thetas[player1]-est_thetas[player2],
#          d_phi = est_phis[hand1]-est_phis[hand2],
#          d_phi2 = atan2(sin(d_phi),cos(d_phi)),
#          exponent = -(d_theta+2*atanh(sinpi(d_phi2))),
#          prob = 1/(1+exp(exponent))) |>
#   group_by(hand1,hand2) |>
#   summarise(expected_payoff=mean(prob),.groups = "drop_last")
# 
# est_payoff_mat <- xtabs(expected_payoff ~ hand1 + hand2, data=post_estimation)
# est_payoff_mat
```

### 4-dice

```{r}
N_games <- 1000
N_ids <- 100
N_opts <- 4
thetas <- rnorm(N_ids)

payoff_mat <- grime_dice_payoffs(N_opts)
grime_dice_matches <- as.data.frame(
  do.call(rbind, 
          replicate(N_games,
                    simulate_rps(thetas,N_opts,payoff_mat,TRUE),
                    simplify = FALSE)))

grime_dice_matches <- grime_dice_matches |> rename(player1="V1",
                             player2="V2",
                             hand1 = "V3",
                             hand2 = "V4",
                             result="V5",
                             true_p="V6")
```

```{r}
elo_sol <- fit_elo_model(N_ids,grime_dice_matches)
elo_pred <- get_predicted_prob(grime_dice_matches,theta=elo_sol$par)
plot_elo(elo_sol,thetas,p_true=grime_dice_matches$true_p,p_pred=elo_pred,w=grime_dice_matches$result)

res_nt <- estimate_nt_elo(N_ids,N_opts,grime_dice_matches,n_start = 10,
                          theta_centering = TRUE,
                          use_atanh = FALSE,
                          scale_by_sd = FALSE,
                          ridge_penalty = TRUE)

estimates <- get_res_estimates(res_nt,N_ids,theta_centering=TRUE)
pred_p <- get_predicted_prob(grime_dice_matches,theta=estimates$theta,phi=estimates$phi)

plot_nt_elo(res_nt$sol,thetas,N_ids,N_opts,theta_centering = TRUE,
            p_true=grime_dice_matches$true_p,p_pred=pred_p,w=grime_dice_matches$result,
            labels=c("Red", "Blue", "Olive", "Yellow", "Magenta"))

est_thetas <- estimates$theta
est_phis <- estimates$phi
post_estimation <- grime_dice_matches |>
  mutate(d_theta = est_thetas[player1]-est_thetas[player2],
         d_phi = est_phis[hand1]-est_phis[hand2],
         d_phi2 = atan2(sinpi(d_phi), cospi(d_phi)) %% (2*pi),
         exponent = -(2*sin(d_phi2)),
         prob = 1/(1+exp(exponent))) |>
  group_by(hand1,hand2) |>
  summarise(expected_payoff=mean(prob),.groups = "drop_last")

payoff_mat
est_payoff_mat <- xtabs(expected_payoff ~ hand1 + hand2, data=post_estimation)
est_payoff_mat


res_nt <- estimate_nt_elo(N_ids,N_opts,grime_dice_matches,n_start = 10,
                          theta_centering=FALSE,use_atanh=TRUE)

estimates <- get_res_estimates(res_nt,N_ids,theta_centering=FALSE)
pred_p <- get_predicted_prob(grime_dice_matches,
                             theta=estimates$theta,phi=estimates$phi,
                             use_atanh = TRUE)

plot_nt_elo(res_nt$sol,thetas,N_ids,N_opts,theta_centering = FALSE,
            p_true=grime_dice_matches$true_p,p_pred=pred_p,w=grime_dice_matches$result,
            labels=c("Red", "Blue", "Olive", "Yellow", "Magenta"))

est_thetas <- estimates$theta
est_phis <- estimates$phi
post_estimation <- grime_dice_matches |>
  mutate(d_theta = est_thetas[player1]-est_thetas[player2],
         d_phi = est_phis[hand1]-est_phis[hand2],
         d_phi2 = atan2(sinpi(d_phi), cospi(d_phi)) %% (2*pi),
         exponent = -(2*atanh(sin(d_phi2))),
         prob = 1/(1+exp(exponent))) |>
  group_by(hand1,hand2) |>
  summarise(expected_payoff=mean(prob),.groups = "drop_last")

payoff_mat
est_payoff_mat <- xtabs(expected_payoff ~ hand1 + hand2, data=post_estimation)
est_payoff_mat

```

### 5-dice

```{r}
N_games <- 1000
N_ids <- 100
N_opts <- 5
thetas <- rnorm(N_ids)

payoff_mat <- grime_dice_payoffs(N_opts)
grime_dice_matches <- as.data.frame(
  do.call(rbind, 
          replicate(N_games,
                    simulate_rps(thetas,N_opts,payoff_mat,TRUE),
                    simplify = FALSE)))

grime_dice_matches <- grime_dice_matches |> rename(player1="V1",
                             player2="V2",
                             hand1 = "V3",
                             hand2 = "V4",
                             result="V5",
                             true_p="V6")
```

```{r}
elo_sol <- fit_elo_model(N_ids,grime_dice_matches)
elo_pred <- get_predicted_prob(grime_dice_matches,theta=elo_sol$par)
plot_elo(elo_sol,thetas,p_true=grime_dice_matches$true_p,p_pred=elo_pred,w=grime_dice_matches$result)

res_nt <- estimate_nt_elo(N_ids,N_opts,grime_dice_matches,n_start = 10,
                          theta_centering = TRUE,
                          use_atanh = FALSE,
                          scale_by_sd = FALSE,
                          ridge_penalty = TRUE)

estimates <- get_res_estimates(res_nt,N_ids,theta_centering=TRUE)
pred_p <- get_predicted_prob(grime_dice_matches,theta=estimates$theta,phi=estimates$phi)

plot_nt_elo(res_nt$sol,thetas,N_ids,N_opts,theta_centering = TRUE,
            p_true=grime_dice_matches$true_p,p_pred=pred_p,w=grime_dice_matches$result,
            labels=c("Red", "Blue", "Olive", "Yellow", "Magenta"))

est_thetas <- estimates$theta
est_phis <- estimates$phi
post_estimation <- grime_dice_matches |>
  mutate(d_theta = est_thetas[player1]-est_thetas[player2],
         d_phi = est_phis[hand1]-est_phis[hand2],
         d_phi2 = atan2(sinpi(d_phi), cospi(d_phi)) %% (2*pi),
         exponent = -(2*sin(d_phi2)),
         prob = 1/(1+exp(exponent))) |>
  group_by(hand1,hand2) |>
  summarise(expected_payoff=mean(prob),.groups = "drop_last")

payoff_mat
est_payoff_mat <- xtabs(expected_payoff ~ hand1 + hand2, data=post_estimation)
est_payoff_mat


res_nt <- estimate_nt_elo(N_ids,N_opts,grime_dice_matches,n_start = 10,
                          theta_centering=FALSE,use_atanh=TRUE)

estimates <- get_res_estimates(res_nt,N_ids,theta_centering=FALSE)
pred_p <- get_predicted_prob(grime_dice_matches,
                             theta=estimates$theta,phi=estimates$phi,
                             use_atanh = TRUE)

plot_nt_elo(res_nt$sol,thetas,N_ids,N_opts,theta_centering = FALSE,
            p_true=grime_dice_matches$true_p,p_pred=pred_p,w=grime_dice_matches$result,
            labels=c("Red", "Blue", "Olive", "Yellow", "Magenta"))

est_thetas <- estimates$theta
est_phis <- estimates$phi
post_estimation <- grime_dice_matches |>
  mutate(d_theta = est_thetas[player1]-est_thetas[player2],
         d_phi = est_phis[hand1]-est_phis[hand2],
         d_phi2 = atan2(sinpi(d_phi), cospi(d_phi)) %% (2*pi),
         exponent = -(2*atanh(sin(d_phi2))),
         prob = 1/(1+exp(exponent))) |>
  group_by(hand1,hand2) |>
  summarise(expected_payoff=mean(prob),.groups = "drop_last")

payoff_mat
est_payoff_mat <- xtabs(expected_payoff ~ hand1 + hand2, data=post_estimation)
est_payoff_mat

```

## Recovering parameters

If the probability of winning is actually a combination of difference in abilities and advantage of the hand played, can we recover these parameters?

```{r}
set.seed(1729)
N_ids <- 100
N_games <- 1000
N_opts <-3

NT_thetas <- rnorm(N_ids)
delta_phi <- 2/N_opts
NT_phis <- runif(N_opts,0,2)
# NT_phis <- seq(from=0,to=(2-delta_phi),length.out=N_opts)

NT_matches <- as.data.frame(do.call(
  rbind, 
  replicate(N_games,simulate_NT_model(NT_thetas,NT_phis,N_opts,TRUE),simplify = FALSE)))

NT_matches <- NT_matches |> rename(player1="V1",
                             player2="V2",
                             hand1 = "V3",
                             hand2 = "V4",
                             result="V5",
                             true_p="V6")
```

```{r}
ggplot(NT_matches,aes(x=true_p,y=result,color=factor(result))) +geom_point(size=1,alpha=0.5) + theme_bw() + labs(x="True Predicted Prob",y="Result",color="Result",title = "True Sim data")

ggplot(NT_matches,aes(x=true_p,fill=factor(result))) +geom_density(alpha=0.5) + theme_bw() + labs(x="True Predicted Prob",y="density",color="Result",title = "True Sim data")
```

```{r}
elo_sol <- fit_elo_model(N_ids,NT_matches)
elo_pred <- get_predicted_prob(NT_matches,theta=elo_sol$par)
plot_elo(elo_sol,NT_thetas,p_true=NT_matches$true_p,p_pred=elo_pred,w=NT_matches$result)

res_nt <- estimate_nt_elo(N_ids,N_opts,NT_matches,n_start = 10,
                          theta_centering = TRUE,
                          use_atanh = FALSE,
                          scale_by_sd = FALSE,
                          ridge_penalty = TRUE)

estimates <- get_res_estimates(res_nt,N_ids,theta_centering=TRUE)
pred_p <- get_predicted_prob(NT_matches,theta=estimates$theta,phi=estimates$phi)

plot_nt_elo(res_nt$sol,NT_thetas,N_ids,N_opts,theta_centering = TRUE,
            p_true=NT_matches$true_p,p_pred=pred_p,w=NT_matches$result,
            labels=c("Rock", "Paper", "Scissors","Spock","Lizard"))

res_nt <- estimate_nt_elo(N_ids,N_opts,NT_matches,n_start = 10,
                          theta_centering=FALSE,use_atanh=TRUE)

estimates <- get_res_estimates(res_nt,N_ids,theta_centering=FALSE)
pred_p <- get_predicted_prob(NT_matches,
                             theta=estimates$theta,phi=estimates$phi,
                             use_atanh = TRUE)

plot_nt_elo(res_nt$sol,NT_thetas,N_ids,N_opts,theta_centering = FALSE,
            p_true=NT_matches$true_p,p_pred=pred_p,w=NT_matches$result,
            labels=c("Rock", "Paper", "Scissors","Spock","Lizard"))
```

```{r}
# elo_sol  <- fit_elo_model(N_ids, NT_matches)
# res_rps  <- estimate_nt_elo(N_ids, N_opts, NT_matches, 10)
# res_rps_exp <- estimate_nt_elo(N_ids, N_opts, NT_matches, 10, theta_centering="fixed", use_atanh=TRUE)
# 
# est_thetas <- as_tibble(elo_sol$par) |> rename(elo_theta='value') |> mutate(id=row_number())
# est_thetas$rps_theta     <- res_rps$sol$par[1:N_ids]
# est_thetas$rps_exp_theta <- c(0, res_rps_exp$sol$par[1:(N_ids-1)])
# est_thetas$true_theta    <- NT_thetas
# 
# est_phis <- as_tibble(c(1:N_opts)) |> rename(hand='value')
# est_phis$rps_phi     <- c(0, res_rps$sol$par[(N_ids+1):(N_ids+N_opts-1)])
# est_phis$rps_exp_phi <- c(0, res_rps_exp$sol$par[N_ids:(N_ids+N_opts-2)])
# est_phis$true_phi    <- NT_phis
# 
# NT_matches_predict <- NT_matches |>
#   left_join(est_thetas, by=join_by(player1==id)) |>
#   left_join(est_thetas, by=join_by(player2==id)) |>
#   left_join(est_phis,   by=join_by(hand1==hand)) |>
#   left_join(est_phis,   by=join_by(hand2==hand)) |>
#   mutate(elo_d_theta     = elo_theta.x - elo_theta.y,
#          rps_d_theta     = rps_theta.x - rps_theta.y,
#          rps_exp_d_theta = rps_exp_theta.x - rps_exp_theta.y,
#          true_d_theta    = true_theta.x - true_theta.y,
#          rps_d_phi       = rps_phi.x - rps_phi.y,
#          true_d_phi      = true_phi.x - true_phi.y,
#          rps_exp_d_phi   = rps_exp_phi.x - rps_exp_phi.y,
#          rps_g_d_phi     = 2*sin(atan2(sinpi(rps_d_phi), cospi(rps_d_phi))),
#          rps_exp_g_d_phi = 2*atanh(sin(atan2(sinpi(rps_exp_d_phi), cospi(rps_exp_d_phi)))),
#          true_g_d_phi    = 2*sin(atan2(sinpi(true_d_phi), cospi(true_d_phi))),
#          p_elo     = plogis(elo_d_theta),
#          p_rps     = plogis(rps_d_theta + rps_g_d_phi),
#          p_rps_exp = plogis(rps_exp_d_theta + rps_exp_g_d_phi),
#          p_true    = plogis(true_d_theta + true_g_d_phi))
# 
# plot_elo(elo_sol, NT_thetas)
# plot_nt_elo(res_rps$sol, NT_thetas, N_ids, N_opts,
#             labels=c("Rock","Paper","Scissors","Spock","Lizard"),
#             p_pred=NT_matches_predict$p_rps,
#             p_true=NT_matches_predict$p_true)
# plot_nt_elo(res_rps_exp$sol, NT_thetas, N_ids, N_opts, theta_offset=1,
#             labels=c("Rock","Paper","Scissors","Spock","Lizard"),
#             p_pred=NT_matches_predict$p_rps_exp,
#             p_true=NT_matches_predict$p_true)
```

```{r}
# ggplot(NT_matches_predict,aes(x=p_elo,fill=factor(result))) +geom_density(alpha=0.5) + theme_bw() + labs(x="ELO Predicted Prob",y="density",color="Result",title = "ELO")
# 
# ggplot(NT_matches_predict,aes(x=p_rps,fill=factor(result))) +geom_density(alpha=0.5) + theme_bw() + labs(x="NT-ELO Predicted Prob",y="density",color="Result",title = "NT-ELO")
# 
# ggplot(NT_matches_predict,aes(x=p_rps_exp,fill=factor(result))) +geom_density(alpha=0.5) + theme_bw() + labs(x="NT-Exp-ELO Predicted Prob",y="density",color="Result",title = "NT-Exp-ELO")
```

### Dice with Handicap

```{r}
dice_3 <- list(c(rep(3,5),6), #Red: 333336
               c(rep(2,3),rep(5,3)), #Blue: 222555
               c(1,rep(4,5))) # Olive: 144444

dice_4 <- list(c(rep(0,2),rep(4,4)), #Red: 004444
               c(rep(3,6)), #Blue: 333333
               c(rep(1,3),rep(5,3)), # Olive: 111555
               c(rep(2,4),rep(6,2))) # Yellow: 222266

dice_5 <- list(c(rep(4,5),9), #Red: 444449
               c(rep(2,3),rep(7,3)), #Blue: 222777
               c(rep(0,1),rep(5,5)), # Olive: 055555
               c(rep(3,4),rep(8,2)), # Yellow: 333388
               c(rep(1,2),rep(6,4))) # Magenta: 116666

```

#### 3-dice

```{r}
N_ids <- 100
N_games <- 1000
N_opts <- 3
dice_thetas <- rnorm(N_ids)

dice_games <- as.data.frame(do.call(rbind,
                      replicate(N_games,simulate_dice_rolls(dice_thetas,dice_3),
                      simplify=FALSE)))

dice_games <- dice_games |> rename(player1="V1",
                             player2="V2",
                             hand1 = "V3",
                             hand2 = "V4",
                             result="V5",
                             true_p="V6")
```

```{r}
elo_sol <- fit_elo_model(N_ids,dice_games)
elo_pred <- get_predicted_prob(dice_games,theta=elo_sol$par)
plot_elo(elo_sol,dice_thetas,p_true=dice_games$true_p,p_pred=elo_pred,w=dice_games$result)

res_nt <- estimate_nt_elo(N_ids,N_opts,dice_games,n_start = 10,
                          theta_centering = TRUE,
                          use_atanh = FALSE,
                          scale_by_sd = FALSE,
                          ridge_penalty = TRUE)

estimates <- get_res_estimates(res_nt,N_ids,theta_centering=TRUE)
pred_p <- get_predicted_prob(dice_games,theta=estimates$theta,phi=estimates$phi)

plot_nt_elo(res_nt$sol,dice_thetas,N_ids,N_opts,theta_centering = TRUE,
            p_true=dice_games$true_p,p_pred=pred_p,w=dice_games$result,
            labels=c("Red", "Blue", "Olive", "Yellow", "Magenta"))


res_nt <- estimate_nt_elo(N_ids,N_opts,dice_games,n_start = 10,
                          theta_centering=FALSE,use_atanh=TRUE)

estimates <- get_res_estimates(res_nt,N_ids,theta_centering=FALSE)
pred_p <- get_predicted_prob(dice_games,
                             theta=estimates$theta,phi=estimates$phi,
                             use_atanh = TRUE)

plot_nt_elo(res_nt$sol,dice_thetas,N_ids,N_opts,theta_centering = FALSE,
            p_true=dice_games$true_p,p_pred=pred_p,w=dice_games$result,
            labels=c("Red", "Blue", "Olive", "Yellow", "Magenta"))

```

```{r}
# elo_sol <- fit_elo_model(N_ids,dice_games)
# plot_elo(elo_sol,dice_thetas)
# 
# res_rps <- estimate_nt_elo(N_ids,3,dice_games,10)
# sol <- res_rps$sol
# plot_nt_elo(sol,dice_thetas,N_ids,3,labels=c("Red","Blue","Olive","Yellow","Magenta"))
# 
# res_rps_exp <- estimate_nt_elo(N_ids,3,dice_games,10,theta_centering="fixed",use_atanh=TRUE)
# sol <- res_rps_exp$sol
# plot_nt_elo(sol,dice_thetas,N_ids,3,theta_offset=1,labels=c("Red","Blue","Olive","Yellow","Magenta"))
```

#### 4-dice

```{r}
N_ids <- 100
N_games <- 1000
N_opts<-4
dice_thetas <- rnorm(N_ids)

dice_games <- as.data.frame(do.call(rbind,
                      replicate(N_games,simulate_dice_rolls(dice_thetas,dice_4),
                      simplify=FALSE)))

dice_games <- dice_games |> rename(player1="V1",
                             player2="V2",
                             hand1 = "V3",
                             hand2 = "V4",
                             result="V5",
                             true_p="V6")
```

```{r}
elo_sol <- fit_elo_model(N_ids,dice_games)
elo_pred <- get_predicted_prob(dice_games,theta=elo_sol$par)
plot_elo(elo_sol,dice_thetas,p_true=dice_games$true_p,p_pred=elo_pred,w=dice_games$result)

res_nt <- estimate_nt_elo(N_ids,N_opts,dice_games,n_start = 10,
                          theta_centering = TRUE,
                          use_atanh = FALSE,
                          scale_by_sd = FALSE,
                          ridge_penalty = TRUE)

estimates <- get_res_estimates(res_nt,N_ids,theta_centering=TRUE)
pred_p <- get_predicted_prob(dice_games,theta=estimates$theta,phi=estimates$phi)

plot_nt_elo(res_nt$sol,dice_thetas,N_ids,N_opts,theta_centering = TRUE,
            p_true=dice_games$true_p,p_pred=pred_p,w=dice_games$result,
            labels=c("Red", "Blue", "Olive", "Yellow", "Magenta"))


res_nt <- estimate_nt_elo(N_ids,N_opts,dice_games,n_start = 10,
                          theta_centering=FALSE,use_atanh=TRUE)

estimates <- get_res_estimates(res_nt,N_ids,theta_centering=FALSE)
pred_p <- get_predicted_prob(dice_games,
                             theta=estimates$theta,phi=estimates$phi,
                             use_atanh = TRUE)

plot_nt_elo(res_nt$sol,dice_thetas,N_ids,N_opts,theta_centering = FALSE,
            p_true=dice_games$true_p,p_pred=pred_p,w=dice_games$result,
            labels=c("Red", "Blue", "Olive", "Yellow", "Magenta"))
```

```{r}
# elo_sol <- fit_elo_model(N_ids,dice_games)
# plot_elo(elo_sol,dice_thetas)
# 
# res_rps <- estimate_nt_elo(N_ids,4,dice_games,10)
# sol <- res_rps$sol
# plot_nt_elo(sol,dice_thetas,N_ids,4,labels=c("Red","Blue","Olive","Yellow","Magenta"))
# 
# res_rps_exp <- estimate_nt_elo(N_ids,4,dice_games,10,theta_centering="fixed",use_atanh=TRUE)
# sol <- res_rps_exp$sol
# plot_nt_elo(sol,dice_thetas,N_ids,4,theta_offset=1,labels=c("Red","Blue","Olive","Yellow","Magenta"))
```

#### 5-dice

```{r}
N_ids <- 100
N_games <- 1000
N_opts <- 5
dice_thetas <- rnorm(N_ids)

dice_games <- as.data.frame(do.call(rbind,
                      replicate(N_games,simulate_dice_rolls(dice_thetas,dice_5),
                      simplify=FALSE)))

dice_games <- dice_games |> rename(player1="V1",
                             player2="V2",
                             hand1 = "V3",
                             hand2 = "V4",
                             result="V5",
                             true_p="V6")

```

```{r}
elo_sol <- fit_elo_model(N_ids,dice_games)
elo_pred <- get_predicted_prob(dice_games,theta=elo_sol$par)
plot_elo(elo_sol,dice_thetas,p_true=dice_games$true_p,p_pred=elo_pred,w=dice_games$result)

res_nt <- estimate_nt_elo(N_ids,N_opts,dice_games,n_start = 10,
                          theta_centering = TRUE,
                          use_atanh = FALSE,
                          scale_by_sd = FALSE,
                          ridge_penalty = TRUE)

estimates <- get_res_estimates(res_nt,N_ids,theta_centering=TRUE)
pred_p <- get_predicted_prob(dice_games,theta=estimates$theta,phi=estimates$phi)

plot_nt_elo(res_nt$sol,dice_thetas,N_ids,N_opts,theta_centering = TRUE,
            p_true=dice_games$true_p,p_pred=pred_p,w=dice_games$result,
            labels=c("Red", "Blue", "Olive", "Yellow", "Magenta"))


res_nt <- estimate_nt_elo(N_ids,N_opts,dice_games,n_start = 10,
                          theta_centering=FALSE,use_atanh=TRUE)

estimates <- get_res_estimates(res_nt,N_ids,theta_centering=FALSE)
pred_p <- get_predicted_prob(dice_games,
                             theta=estimates$theta,phi=estimates$phi,
                             use_atanh = TRUE)

plot_nt_elo(res_nt$sol,dice_thetas,N_ids,N_opts,theta_centering = FALSE,
            p_true=dice_games$true_p,p_pred=pred_p,w=dice_games$result,
            labels=c("Red", "Blue", "Olive", "Yellow", "Magenta"))
```

```{r}
# elo_sol <- fit_elo_model(N_ids,dice_games)
# plot_elo(elo_sol,dice_thetas)
# 
# res_rps <- estimate_nt_elo(N_ids,5,dice_games,10)
# sol <- res_rps$sol
# plot_nt_elo(sol,dice_thetas,N_ids,5,labels=c("Red","Blue","Olive","Yellow","Magenta"))
# 
# res_rps_exp <- estimate_nt_elo(N_ids,5,dice_games,10,theta_centering="fixed",use_atanh=TRUE)
# sol <- res_rps_exp$sol
# plot_nt_elo(sol,dice_thetas,N_ids,5,theta_offset=1,labels=c("Red","Blue","Olive","Yellow","Magenta"))
```

## Starcraft II

```{python}
# Python code for getting Starcraft data
# 
# import requests
# import numpy as np
# import pandas as pd
# import json
# 
# key = "KzTkOG5PJS1HC6uFBtt6"
# next_url = f"/api/v1/match/?apikey={key}&format=json&limit=0&date__gte=2026-01-01&order_by=date&game=LotV"
# api_url = f"https://aligulac.com"
# games = []
# while next_url:
#     print(next_url)
#     res = requests.get(api_url+next_url).json()
#     games.extend(res.get('objects'))
#     next_url = res.get('meta').get('next')
# lotv = pd.DataFrame(games).drop(columns=['event','eventobj','offline','resource_uri','treated']).dropna()
# lotv['pla'] = lotv.pla.apply(lambda x: x['id'])
# lotv['plb'] = lotv.plb.apply(lambda x: x['id'])
# lotv['rta'] = lotv.rta.apply(lambda x: x['rating'])
# lotv['rtb'] = lotv.rtb.apply(lambda x: x['rating'])
# lotv['win'] = [int(a>b) for a,b in zip(lotv['sca'],lotv['scb'])]
# lotv = lotv.rename(columns={
#     'pla':'player1',
#     'plb':'player2',
#     'rca':'hand1',
#     'rcb':'hand2',
#     'win':'result',
# })
# mapping = {
#     "T":1,
#     "Z":2,
#     "P":3,
#     "R":4
# }
# lotv.hand1 = lotv.hand1.map(mapping)
# lotv.hand2 = lotv.hand2.map(mapping)
# # lotv.to_csv('~/Downloads/lotv_2026.csv',index=False)
```

```{r,eval=FALSE}
starcraft <- read.csv("lotv_2026.csv")

starcraft <- starcraft |> filter(hand1!=4,hand2!=4,sca!=scb) 

N_ids_old <- 1e6
N_ids <- 0
while (N_ids != N_ids_old){
  N_ids_old <- N_ids
  player_counts <- bind_rows(
  starcraft |> select(player = player1),
  starcraft |> select(player = player2)
  ) |>
  count(player, name = "games_played")
  
  # Keep only players with at least 10 games
  valid_players <- player_counts |>
    filter(games_played >= 10) |>
    pull(player)
  
  # Filter matches where BOTH players are valid
  starcraft_filtered <- starcraft |>
    filter(player1 %in% valid_players,
           player2 %in% valid_players)
  
  ids <- unique(sort(c(starcraft_filtered$player1,starcraft_filtered$player2)))
  N_ids <- length(ids)
  N_opts <- length(unique(c(starcraft_filtered$hand1,starcraft_filtered$hand2)))
  starcraft <- starcraft_filtered
  
}

starcraft_filtered <- starcraft_filtered|> mutate(player1 = match(player1,ids),
                                player2 = match(player2,ids)) 

```

```{r}
starcraft_filtered <- starcraft_filtered |> mutate(p_rt = plogis(rta-rtb),
                                                   res_delta = sca - scb)
ggplot(starcraft_filtered,aes(x=p_rt,y=res_delta,color=factor(result))) + geom_point() + theme_bw() + labs(x="Prob. from given rating and ELO model",y="Result",color="Result")

ggplot(starcraft_filtered,aes(x=p_rt,fill=factor(result))) + geom_density(alpha=0.5) + theme_bw() + labs(x="Prob. from given rating and ELO model",y="density",fill="Result")

```

```{r,eval=FALSE}
avg_player_ratings <- rbind(starcraft_filtered |> select(player1,rta) |> group_by(player1) |> summarise(theta=mean(rta)) |> rename(player=player1),
      starcraft_filtered |> select(player2,rtb) |> group_by(player2) |> summarise(theta=mean(rtb))|>rename(player=player2)) |> group_by(player) |> summarise(theta=mean(theta))
```

```{r,eval=FALSE}
res_nt_elo <- estimate_nt_elo(N_ids, N_opts, starcraft_filtered,15)
sol <- res_nt_elo$sol
plot_nt_elo(sol,avg_player_ratings$theta,N_ids,N_opts,labels=c("T", "Z", "P"))
```

```{r}
## with sd_adjustment
res_nt_elo_sd <- estimate_nt_elo(N_ids, N_opts, starcraft_filtered,15)
sol <- res_nt_elo_sd$sol
plot_nt_elo(sol,avg_player_ratings$theta,N_ids,N_opts,labels=c("T", "Z", "P"))
```

```{r,eval=FALSE}
starcraft_matchup_summary <- starcraft_filtered |> group_by(hand1,hand2) |> summarise(m=mean(result),.groups = "drop_last")

xtabs(m ~ hand1 + hand2, data=starcraft_matchup_summary)
```

```{r,eval=FALSE}
res_elo <- fit_elo_model(N_ids, starcraft_filtered)
plot_elo(res,avg_player_ratings$theta)
```

```{r}
est_thetas <- as_tibble(res_elo$par) |> rename(elo_theta='value') |> mutate(id=row_number())
est_thetas$rps_theta <- c(0,res_nt_elo$sol$par[1:(N_ids-1)])
est_thetas$rps_theta_sd_adj <- c(0,res_nt_elo_sd$sol$par[1:(N_ids-1)])
est_thetas$true_theta <- avg_player_ratings$theta
  
est_phis <- as_tibble(c(1:N_opts)) |> rename(hand='value')
est_phis$rps_phi <- c(0,res_nt_elo$sol$par[N_ids:(N_ids + N_opts -2)])
est_phis$rps_phi_sd_adj <- c(0,res_nt_elo_sd$sol$par[N_ids:(N_ids + N_opts -2)])

starcraft_filtered_predict <- starcraft_filtered |> 
  left_join(est_thetas,by=join_by(player1==id)) |> 
  left_join(est_thetas,by=join_by(player2==id)) |>
  left_join(est_phis,by=join_by(hand1==hand)) |>
  left_join(est_phis,by=join_by(hand2==hand)) |>
  mutate(elo_d_theta = elo_theta.x - elo_theta.y,
         rps_d_theta = rps_theta.x - rps_theta.y,
         rps_sd_d_theta = rps_theta_sd_adj.x - rps_theta_sd_adj.y,
         true_d_theta = true_theta.x -true_theta.y,
         rps_d_phi = rps_phi.x - rps_phi.y,
         rps_sd_d_phi = rps_phi_sd_adj.x - rps_phi_sd_adj.y,
         rps_g_d_phi = 2*sin(atan2(sinpi(rps_d_phi),cospi(rps_d_phi))),
         rps_sd_g_d_phi = 2*sin(atan2(sinpi(rps_sd_d_phi),cospi(rps_sd_d_phi))),
         p_elo = plogis(elo_d_theta),
         p_rps = plogis(rps_d_theta+rps_g_d_phi),
         p_rps_sd = plogis(rps_sd_d_theta+rps_sd_g_d_phi),
         p_true = plogis(true_d_theta)
         )
```

```{r}

p1 <- ggplot(starcraft_filtered_predict,aes(x=p_elo,y=result,color=factor(result))) +geom_point(size=1,alpha=0.5) + theme_bw() + labs(x="ELO Predicted Prob",y="Result",color="Result",title = "ELO Model")

p2 <- ggplot(starcraft_filtered_predict,aes(x=p_rps,y=result,color=factor(result))) +geom_point(size=1,alpha=0.5) + theme_bw() + labs(x="NT-ELO Predicted Prob",y="Result",color="Result",title = "NT-ELO Model")

p3 <- ggplot(starcraft_filtered_predict,aes(x=p_rps_sd,y=result,color=factor(result))) +geom_point(size=1,alpha=0.5) + theme_bw() + labs(x="NT-ELO Predicted Prob",y="Result",color="Result",title = "NT-ELO Model (sd-adjusted)")

p1 | p2 | p3

```

```{r}
ggplot(starcraft_filtered_predict,aes(x=p_elo,fill=factor(result))) +geom_density(alpha=0.5) + theme_bw() + labs(x="ELO Predicted Prob",y="density",color="Result",title = "ELO")

ggplot(starcraft_filtered_predict,aes(x=p_rps,fill=factor(result))) +geom_density(alpha=0.5) + theme_bw() + labs(x="NT-ELO Predicted Prob",y="density",color="Result",title = "NT-ELO")

ggplot(starcraft_filtered_predict,aes(x=p_rps_sd,fill=factor(result))) +geom_density(alpha=0.5) + theme_bw() + labs(x="NT-ELO Predicted Prob",y="density",color="Result",title = "NT-ELO")

```

```{r}
mean(abs(starcraft_filtered_predict$p_elo - starcraft_filtered_predict$result))
mean(abs(starcraft_filtered_predict$p_rt - starcraft_filtered_predict$result))
mean(abs(starcraft_filtered_predict$p_rps - starcraft_filtered_predict$result))
mean(abs(starcraft_filtered_predict$p_rps_sd - starcraft_filtered_predict$result))
mean(abs(starcraft_filtered_predict$p_true - starcraft_filtered_predict$result))
```

```{r}
get_acc <- function(col='p_elo',t=0.5){
  n_correct = sum((starcraft_filtered_predict[[col]] >= t)*(starcraft_filtered_predict$result) + (starcraft_filtered_predict[[col]] < t)*(1-starcraft_filtered_predict$result))
  acc = n_correct/nrow(starcraft_filtered_predict)  
  return(acc)
}

max_ps <- rep(0,5)
best_ts <- rep(0,5)
p_accs <- rep(0,5)
for(t in seq(0.45,0.55,0.01)){
  p_accs[1] <- get_acc('p_rt',t)
  p_accs[2] <- get_acc('p_elo',t)
  p_accs[3] <- get_acc('p_rps',t)
  p_accs[4] <- get_acc('p_rps_sd',t)
  p_accs[5] <- get_acc('p_true',t)
  
  for(i in c(1:5)){
    if(p_accs[i] >= max_ps[i]){
      max_ps[i] <- p_accs[i]
      best_ts[i] <- t
    }
  }
  
}
max_ps
best_ts


```